\documentclass[12pt]{article}
\usepackage{fontspec}
\usepackage[draft]{microtype}
\usepackage{xltxtra}
\usepackage[margin=1in]{geometry}
\usepackage{multicol}
\setmainfont[Numbers=OldStyle,Scale=0.9]{Linux Libertine O}

\begin{document}

\textbf{Report of the Defense Science Board Task Force on}

\textbf{MILITARY\\ SOFTWARE}

SEPTEMBER 1987

Office of the Under Secretary of Defense for Acquisition\\
Washington, D.C. 20301

\newpage

% Memo cover page for the report

OFFICE OF THE SECRETARY OF DEFENSE
WASHINGTON, D.C. 20301-3140

DEFENSE SCIENCE BOARD

October 7, 1987

MEMORANDUM FOR UNDER SECRETARY OF DEFENSE FOR ACQUISITION

SUBJECT: Report of the Defense Science Board Task Force on Military Software

I am pleased to forward the final report of the Defense Sclonce Board Task
Force on software. This Task Force, chaired by Dr. Frederick Brooks, addressed
the managerial and technical changes needed to improve the software
acquisition process within the DoD.

Although the report has been delayed and the technical assessment of the STARS
program is dated, the report is still a valuable document and should be widely
distributed. The report's recommendations call for hard management thinking on
how to accomplish the iterative setting of requirements within the existing
acquisition structure. The report also concludes that software productivity
improvements will be achieved mainly through management initiatives within the
contracting process as opposed to technical "magic."

The Task Force's assessments on the Ada language initiative and the Software
Engineerine Institute are also valuable.  I recommend that the report be
distributed to the offices within OSD, OJCS, the Service staffs that deal with
software technology and acquisition, and to the appropriate Defense related
industries and organizations.

Charles A. Fowler
Chairman

Attachment

% Memo from Dr. Brooks submitting the report

OFFICE OF THE SECRETARY OF DEFENSE
WASHINGTON, D.C. 20301-3140

DEFENSE SCIENCE BOARD

July 1, 1987

MEMORANDUM FOR CHAIRMAN,

SUBJECT: Report of the Defense Science Board Task Force on Military Software

I enclose the final report of the Defense Science Board Task Force on Military
Software.  Our major findings and recommendations may be found in the
Executive Summary.  It has been a delight to serve with such an expert and
diligent group of people on this Task Force. We would all like to express our
gratitude to the many people who briefed us, and from those supporting this
effort, especially Lieutenant Colonel Susan Swift, USAF.

I regret that my personal difficulties have so long delayed the report's final
submission.  As a result of these delays, our technical assessment of STARS is
dated, although I have been recently briefed on its progress.  The program has
acquired considerable focus under Colonel Green, and is moving forward.  Our
recommendation for its organizational transfer remains unchanged, however, by
these developments.

Some of our recommendations have been implemented already.  The others remain
current and valid.

Frederick P. Brooks, Jr.
Chairman
Defense Science Board
Task Force on Military Software

Attachment

% title
Report of the Task Force on Military Software
Defense Science Board

1. Executive Summary

Many previous studies have provided an abundance of valid conclusions and
detailed recommendations. Most remain unimplemented. If the military software
problem is real, it is not perceived as urgent. We do not attempt to prove
that it is; we do recommend how to attack it if one wants to.

We do not see any single technological development in the next decade that
promises ten-fold improvement in software productivity, reliability, and
timeliness. There are several technical developments under way which together
can be expected to yield one order of magnitude, but not two. Few fields have
so large a gap between best current practice and average current practice; we
concur with the priorities that DoD has given tc upgrading average practice by
more vigorous technology transfer.

Current DoD initiatives in software technology and methodology include the Ada
effort, the STARS program, DARPA's Strategic Computing Initiative, the
Software Engineering Institute, and a planned program in the Strategic Defense
Initiative. These five initiatives are uncoordinated. We recommend that the
Undersecretary of Defense (Acquisition) establish a formal program
coordination mechanism for them, (Rec. \#2).

The big problems are not technical. In spite of the substantial technical
development needed in requirements-setting, metrics and measures, tools, etc.,
the Task Force is convinced that today's major problems with military software
development are not technical problems, but management problems. Hence we call
for no new initiatives in the development of the technology, some modest shift
of focus in the technology efforts under way, but major re-examination and
change of attitudes, policies, and practices concerning software acquisition.

\subsection*{STARS}

The DoD program for Software Technology for Adaptable, Reliable Systems,
STARS, has made little progress in recent years and has had vague and
ill-focused plans for the future. Service support and enthusiasm is lacking.
Yet it is very important that such a project-independent methodology
development effort proceed. We recommend that the STARS Joint Program Office
be moved from the Office of the Secretary of Defense to the USAF Electronic
Systems Division. (Rec. \#1) We recommend that a, general officer be given
responsibility for STARS, the Ada Joint Program Office, and Software
Engineering Tnstitute (whose contracting office is already in ESD). Deputies
from the other Services should be appointed.

\subsection*{Ada}

It is very important for DoD to have a standard programming language; Ada is
by far the strongest candidate in sight. The 1983 mandate for Ada was
technically premature. DoD commitment to Ada since that time has been weak.
The state of Ada compiling technology is now such that it is time to commit
vigorously and wholeheartedly.  The directives 3405.1 and 3405.2 are right
first steps - management follow-through on enforcement and support is now
essential.

Ada embodies and facilitates a set of new approaches to building software,
generally known as "modern software practices." We expect these practices,
rather than yet another programming language, to make a real difference in
software robustness, reusability, adaptability, and maintenance. Ada is not
the only conceivtble vehicle for such practices, but if is here, it has been
tailored for the embedded software problem, and multiple compilers have- been
validated. We recommend against further waiting, language tuning, or
subsetting.

Achieving the benefits of modern programming practices requires the development of
unified programming environments. This work must continue to be pushed forward.
Few program managers will want to take on the headaches of being first user of a new
tool, yet it is essential that all major new programs be committed to that tool if it is to
be effective. Only top-level DoD commitment and mandate can make that happen.
We commend AJPO for its technical success in establishing the language definition and
language validation procedures. We recommend that it be moved from OSD to a unified
software joint program office in the USAF Electronic Systems Command (Recs. \#6,7).

\subsection*{Acquisition}

\textbf{Mileu}. The civilian software market has exploded in the past decade, so that the total
civilian market for purchased software, not counting in-house-built application software, is
now more than ten times larger than the DoD market. This requires a radical update in
much DoD thinking. Some implications:

\begin{enumerate}
\item DoD can nf longer create de facto standards and enforce them on the civilian market,
as it was able to do with COBOL.

\item DoD must not diverge too far from whatever the civilian market is doing in programming methodology, else it will have to support its own methodology by itself, with little
resource or training commitment from others. (The same thing is true of processor
architectures.)

\item DoD should be aggressively looking for opportunities to buy, in the civilian market,
tools, methods, environments, and application software. Whenever it can use these
instead of custom-built software, it gets big gains in timeliness, cost, reliability,
completeness of documentation, and training. But today's acquisition regulations
and procedures are all heavily biased in favor of developing custom-built software
for individual programs.
\end{enumerate}

\textbf{Life-cycle model}. DoD Directive 5000.29 and STD 2167 codify the best 1975 thinking
about software, including a so-called "waterfall" model calling for formal specification, then
request for bids, then contracting, delivery, installation, and maintenance. In the decade
since the waterfall model was developed, our discipline has come to recognize that setting
the requirements is the most difficult and crucial part of the software building process, and
bne that requires iteration between the designers and users. In best modern practice, the
ea-.y specification is embodied in a prototype, which the intended users can themselves
drive in order to see the consequences of their imaginings. Then, as the design effort
begins to yield data on the cost and schedule consequences of particular specifications, the
designers and the nsers revise the specifications.

Directive 5000.29 not only does not encourage this best znodern practice, it essentially
forbids it. We recommend that it be revised immediately to mandate and facilitate early
prototyping before the baseline specifications are established '(Rec. \#23).

DoD-STD-2167 likewise needs a radical overhaul to reflect best modern practice. Draft
DoD-STD-2167A is a step, but it does not go nearly far enough. As drafted, it continues to
reinforce exactly the document-driven, specify-then-build approach that lies at the heart
of so many DoD software problems.

For major new software builds, we recommend that competitive level-of-effort contracts
be routinely let for determining specificatlons and preparing an early prototype (Rec. \#26).
The work of specification is so crucial, and yet its fraction of total cost is so small, that
we believe duplication in this phase will save money in total program cost, and surely save
time. After a converged-specification has been prepared and validated by prototyping, a
single competitively-bid contract for construction is appropriate.

\textbf{Incentives.} Defense procurement procedures discourage contractor investment in the
development of new software methodology. Any such contractor investment made today
promises low return. We recommend that the DoD rights-in-data policy be revised to
distinguish software rights from other rights, and that the policy as it applies to software
be designed to encourage contractor investment, both with private and IRD funds, in tools,
methods, and programming environments (Recs. \#17-22).

Similarly, today's policies actively discourage the reuse of software modules from
one system in another. We recommend a variety of policy changes, each designed to
encourage reuse, and indeed, the establishment of a public market in reusable software
parts (Recs. \#29-33).

\subsection*{Personnel}

It appears that the number of software-qualified military officers has been
essentially constant over the past decade, despite exponential growth in
software. Many studies have recommended actions that need to be taken re
training, specialty codes, career paths, etc., to address the shortage of
uniformed specialists. Some of these have been taken.  Nevertheless, the
number has not increased.

We doubt that it will. The powerful civilian demand for such persoas will, we expect,
continue to drain thpm away from the Services as fast as they reach first retairement age,
or before. % the PDF is missing the word prior to 'before', I believe it is 'or'

Therefore we recommend that the Services now assume that there will not be more
such people, and concentrate effort on how best to use those they have (Rec. \#34). The
application-knowledgeable, technically skilled leaders are the military's limiting resource
in using today's computer technology.

We observe that in the best military software programs, the number of customer
software people engaged in the acquisition and program oversight approximates 10% of the
number of contractor personnel. This number does not seem too high. Few program offices
are staffed so well, however, largely due to the shortage of qualified people. Meanwhile
one observes some substantial software-building efforts under way within the Services,
usually done by a combination of civilian and uniformed personnel, generally managed by
software-qualified officers. This is a second-best use of the available specialist officers.

We recommend phasing out this practice and concentrating the available knowledgeable
officers on acquisition (Rec. \#35). We see no other way that the exponential growth in
needed military software can be met.

%2. Introduction
2.1 The Charge to the Task Force
Abbreviated Terms of Reference, (Appendix A2 contains the faill text).
A. Assess and unify various recent studies.
B. Examine why software costs are high.
C. Assess STARS for military software; discuss the priority of its components.
D. Recommend how to enlist industry, Service, and university efforts in a productivity
thrust.
E. Assess STARS, etc., for U.S. international competitiveness.
F. Recommend how to apply R\&D funds to get the most increase in military software
capability.
G. Recommend how to implement an incremental and evolutionary approach to (F).
H. Assess the wisdom of the Ad\& plan, especially in view of "Fourth-Generation' languages.
What the Task Force Did Not Address
Problem Seriousness Sizing. It would be presumptuous, and appear to be self-serving
as well, for this Task Force to tell the Service commanders and the DoD civilian authorities
that your mission-critical software problem ranks high on your present or future criticalproblem list. Other studies have sized the cost and recounted software-caused delays and
system malfunctions. Your own experience will have to put this problem into proper
perspective among all your difficulties.
What the Task Force is qualified to do for you is to

V

e characterize software, its problems, and its technology
£

identify trends that will, in the course of time, make today's problems worse or
better,

* suggest actions to address today's problems and avert tomorrow's calamities.
Non-Mission-Critical Software.
The Task Force largely limited itself to missioncritical systems, those wherein military software most differs from civilian-market software.
Our recommendations wi'.h respect to procurement, however, apply to all DoD acquisition
of software, In Section 6 we categorize DoD software according to the degree to which it

%nrug

r

r

w I

.

.

-

.

.

-----

--

must be non-standard because of its military function.
Service-Specific Personrel Problems. We did not address Service-specific personnel
and skills problems. Thesm have been adequately addrdissed in earlier studies. The careerpath and skills-retention problems continue to be very real in all the Services.
SE1. We did not review the Software Engineering Institute, other than to hear a briefing
on its objectives. It, was in the process of being established and finding a permanent
director during our study; any review would have been premature.
SDI. The same was true of the SDI plan for developing software methodology. At the
tit~ie we were briefed by the SDI office, there was no plan to review.
SCI. We hadi only one briefing on the DARPA soft-ware methodology efforts encompassed
within the 'Strategic Computing Initiative. These efforts are properly aimed at producing
results a decade hence. The approaches are sufficiently bold that little in the way of
directly ?.pplic able short- and mid-term results can be expected.
New 10echnological Initiatives. We do not recommend any new initiatives or funding
for n'ýw specific research or technology-development programs. We support the recent
technological initiatives, but today's major unaddressed problems are not technical, but
managerial.
2.2 Military Software

Role.

Software plays a major role in today's weapon systems. The "smarts" of smart

weapons are provided by software. Software is crucial to intelligence, communications,
command, and control. Software enables computerized systemns for logistics, personnel,
and finance. The chief "military software problem7 is that we cannot get enough of it,
soon enough, reliable enough, and cheap enough to meet the demands of weapon systems
desgnrsand users. Software provides a major Lomponent of U.S. war-fighting capability.
Growth. DoD software-intensive systems have grown exponentially, reaching an annual
software expenditure level in mission-critical computer systems of about \$9 billion in 1985,

with projections of \$30 billion annually by 1990 [Taft, 1985).

This continuing growth

has strained the ability of the DoD to manage their development. Because software
controls system function, deficiencies in software development affect over-all weapon system
performance and cost quite out of proportion to the software cost itself.

Like Civilian Software.

Military software is fundamentally like advanced civilian

software, only more so. That is, the properties of real-time operational software in civilian
banking, airline reservations, or process control, are the same as those of weapon-csystem
software. Big civilian database and file systems look essentially like the military logistics,
6
6%

%finance, and persoDnel software. In the operation of a ship or a base, one finds many small
computers whose tasks are essentially the x.n-e as those in civilian businesses.
Only More So.
Mission-critical military software is more universally real-time,
communications-oriented, and resource-constrained thaw its civilian counterparts. At any
given time, the demands of weapon systems stress the state of the software art more
severely than do rivilian demands.
Timeliness and Reliability. Although the cost of military software is commonly seen
as the major problem, and is emphasized in our Terms of Reference, both previous studies
and our briefers suggest that software timeliness and reliability are even more critical
problems today.
Software development cycles are long, relatively unpredictable, and come at the end of
total weapon system development. Thus they frequently encounter delays, delays usually
on the critical path to operational capability. It also takes too long to adapt running
software to changing hardware or operational requirements.
Software reliability is equally of concern. Since operational software i. complex, it
usually contains design flaws, and these are hard to find and often painful in effect.
Requirements-Setting Is The Hardest Part.
As is true for complex hardware
systems, the hardest part of the software task is the setting of the exact requirements,
including numbers for size and performance, and including the relative priorities of different
requirements in the designers' inevitable trade-offs.
We have no technology and only poor methodologies for establishing such requirements.
There are not even good ways in common use for even stating detailed requirements and
trade-off priorities. Misjudgements in requirements badly hurt effectiveness, cost, and
schedule. Such misjudgements abound. Most common is the specification of over-rich
function, whose bad effects on size and performance become wvident only late in the design
cycle. Another common error is the mis-imagination of how user interfaces should work.
In our view the difficulty is fundamental. We believe that users cannot, with any
amount of effort and wisdom, accurately describe the operational requirements for a substantial software system without testing by real operato;s in an o~perational environment,
and iteration on the specification. The systems built today age just too complex for
the mind of man to foresee all the ramifications purely by the exercise of the analytic
imagination.
This inherent difficuiL is unnecessarily compounded in DoD by the presence of too
many intermediaries be" veen the ultimate user and the software specifier.
The Big Problems Are Not Technical. In spite of the substantial technical development needed in requirements-setting, metrics and measures, tools, etc., the Task Force
is convinced that today's major problems with military software development are not
technical problems, but management problems. Hence we call for no new initiatives in the
7

%development of the technology, some modest shift of focus in the technology efforts under
way, but major re-examination and change of attitudes, policies, and practices concerning
softwa~re acquisition.

2.3 Why Is Software Technology Developing So Sflowly?

*

Participants and observers in the computer game often marvel that the software te~zh~no4ogy develops so slowly, especially in comparison with. computer hardware technology.
In our Terms of R~eference we are charged with examining the underlying nature of the
software process so as to explain high costs and slow development.

Hardware Technology Is So Fast.
The remarkable fact is not Ahe slow rate of development of computer software technology, but the fast rate of hardware technology, a fact especially striking to those of us
who do both. Today's hardware offers at least a 10,000-fold gain in price-performance over
that of 30 years ago, and one can choose at least 1000-fold of that gain in either price
or performance! No other technology has come even close to that rate of development.
It reflects the shift of computer hardware from ar, asmembly technology to a process
technology.

Software Is Labor-Intensive.
Software development is and always will be a labor-intensive technology. The work
and the time is all in development, not production. Development ib always labor-intensive.
Moreover, in the ultimate, one is developing conceptual structures, and although our
machines can do the dog-work and can help us keep track of our edifices, concept
development is the quintessentially human activity.

The Essence Is Designing Intricate Conceptual Structures Rigorously.
In Appendix AS, we analyze the software task. We argue that its essence is the
designing of intricate conceptual structures, rigorously and correctly. The part of software
development that will not go away is the craftiUng of these conceptual structures; the part
that can go away is the labor of expressing them. The task is made more difficult by three
other properties of software products: (1) the necessity for them to conform to complex
environmental, hardware, and user interfaces; (2) the necessity for them to change as their
interfaces change; (3) and the invisibility of the structures themselves.
We believe a significant fraction of software development effort today is expended on
this essential 1Nbor, rather than on the task of expressing the designs.

%The Removal of ]Expression Difficulties Has Brought Much of the Past Gain.
The essential labor itself has not always taken most of the effort. Much of the work was
formerly spent on non-essential, incidental difficulties in the expression of the conceptual
structures. The three big breakthroughs in software methodology each have consisted of
removing one of these incidental difficulties.
First was the avwkwardneis of machine language. High-level languages removed this
difficulty and improved productivi•yv ten-fold.
Second was the loss of mental continuity occasioned by slow turn-around batch
compilation and execution. Time-sharing removed this difficulty, improving productivity
2-5 times.
Third was the utter incompatibility of files, formats, and interfaces among various
software tools. Integrated programming environments such as Unix and Interlisp overcame
this difficulty, again doubling (or better) productivity.
What's In the Cards?
There are stll non-essential expression difficulties, but they do not account for most
of the development effort in modern software shops. Future methodological imprcvements
will have to attack the essence - conceptual design itself.
Examination of the most promising technological developments shows no single technique that can be expected to yield as much as a 10-fold improvement in productivity,
timeliness, and robustness in the next ten years.
On the other hand, all of the various technological developments on the horizon
together should easily yield a 10-fold improvement in the next decade. It is not likely
that all those developments together will yield a 100-fold improvement.

2.4 Current Software Trends
Mi

Five developments in the past decade have revolutionized the software scene. DoD
software practices evolved in the '63's and '70's, and they neither take into account nor
utilize these advances.

The Microcomputer Revolution and the Personal Computer
The microcomputer, both as a component, and by its incorporation into personal
computers, has totally changed the computer field and the software field. Every procedure
for computer acquisition, etc., must now define a floor in machine size below which it is not

9

%applicable, and machines below the flior should be treated as commodities, components,
and spare parts. (Not all procedures have yet been so revised.)
Obviously software standards such as the Ada mandate must have such a floor as well.
The constraints on embedded microprocessors are su..h that their software often must be
in machine language. We do not address microprocessor software.
We likewise do not deal with the software problems of personal computers. DoD,
like every large enterprise, needs some standards as to h~w such machines are to be
supplied, how they will be equipped with standard-function programs, and how they are to
ir.'Prchange information. Such standardization should be minimal and light-handed. We
Qld not recommend that the Ada mandate cover personal computers.
America's greatest comparative military advantage is the individual initiative and
ingenuity of our Service people. We are therefore greatly encouraged to see the Services
making personal computers readily available to individual units so that individuals can
solve their own simple computing problems their own way. A personal computer and an
electronic spread sheet make a powerful combination, sufficient for countless tasks.*
A Mass Market for Software
The personal computer revolution has explosively fueled the development of a mass
market for third-party developed software. This is the most important development in the
software field in our time.
Each of several computer architectures (the properties of a computer that determine
what programs it will run) define a market. The biggest are those for IBM PC-compatibles,
Apple-compatibles, Macintosh, DEC VAX-compatibles, Unix-compatibles, and IBM 370compatibles. For each of these markets literally hundreds of packages are available, covering
an immense spectrum of functions and costing from a few dollars to a few hundred
thousand. The markets are fiercely competitive.
Technology for Software Modularization and Reuse
Techniques for designing software in little modules, for defining the module interfaces
p-ecisely, and for using common file formats have come into standard use during the
decade. These methods, the backbone of so-called "modern programming practices",
radically improve the structure and adaptability of large programs. They a~so define
modules, whose reuse often costs one-tenth as much as writing another module to do the
same function. Reuse is also much quicker, and it yields better tested, more reliable code.
taDrs. Jones and Brooks had the opportunity to observe a Blue-Flag simulated Air
Force-Army-Marine tactical exercise. We saw a number and variety of personal computers
that have been integrated effectively into unit operational functions; we were pleased to
see a light dependence on massive computer systems.

10

%The Ada programming language is designed to make such modularization natural,
and to provide very powerful facilities- for linking modules. Integrat.-d programming
environments, such as Unix, provide the same kind of facility at another level, that of
the shell-script linking whole programs together.

Rapid Prototyping and Iterative Development
As people have recognized that the requirements, and especially the user interface,
require iterative development, %%-.th interspersed testing by users, there has developed a
technology for constructing "rapid" prototypes. Such a prototype typically executes the
mo in-line function of its type, but not the countless exceptions that make programming
costly. It usually does not have complete error-handling, restart, or help facilities. The
prototype is often built using a lash-up of handy components that swap performance for
rapid interconnect ability. It is usually run on a computer that is bigger and faster than
the target machine.
Commercial packages enable one to prototype graphics interfaces, for example, so that
user testing can be done quite early in the development.

ProfessionalHumility and Evolutionary Development
Experience with confidently specifying and painfully building mammoths has shown it
to be simplest, safest, and even fastest to develop a complex software system by building
a mninimaal version, putting it into actual use, and then adding function, enhancing speed,
reducing size, etL., according to the priorities that emerge from the actual use. Software
engineers must recognize that we cannot specify mam~moths right the first time. In practice,
Version 2 is usually under development before Version 1 is delivered, so Version 3 may be
the first to be affected by actual experience.
This procedure speeds ffirst delivery. It also provides for the iterative setting of
requirements. It minimizes '%-he specification of heavy function whose performance penalties
have not yet been weighed. It tends to concentrate development effort where it will make
the most difference. Seeing the minimal version run does wonders for the morale of the
development team and substantially boosts their communication as to further development.
Evolutionary development is best technically, and it saves time and money. It plays

E

havoc with the customary forms of competitive procurement, however, and they with it.

0

Creativity in acquisition is now needed.
2.5 Current DoD Programs on Software Technology
Besides some substantial efforts in individual Service laboratories, DoD has under way
five programs aimed at enhancing software methodology:

%STARS. The program for Software Technology for Adaptable, Reliable Systems, managed
by the Undersecretary of Defense (Acquisition), was started in ii80 to address all aspects
of modern software methodology as applied especially to mission-critical computer systems.
ADA.

The Ada program, managed by the Ada Joint Program Office under the Under.

secretary of Defense (Acquisition), was started in 1975 to define and develop a standard
high-level language suitable for embedded computer systems.
Founded in 1964, the SEI mission is focused on
Software Engineering Institute.
technology transfer - bringing the best modern methodology into actual practice in the
Services and among DoD contractors. The SEI is operated by Carnegie-Mellon University
un,'er contract from the USAF Electronic Systems Command.
Strategic Computing Initiative. A software component under the DARPA Strategic
Computing T-nitiative is aimed at developing radically new methods and tools, especially
those based or. expert systems and other artificial intelligence techniques. The program
aims at results a decade ahead of modern practice.
A software component under the Strategic Defense
Strategic Defense Initiative.
Initiative is aimed at providing methodological advances for the building of the massive
distributed, ultra-high-performance software system demanded by the SDI.

2.6 Recent Previous Studies
The Task Force reviewed the available studies done since 1982, starting with the
monumental 1982 Druffel study, which in turn summarized the results of 26 previous
studies. Appendix A4 lists the recent ones.
To a surprising degree, the conclusions of these studies agree with each other and
remain valid; the recommendations continue to be wise. The chairmen of the several study
groups briefed us. All had one message: very little action, has been taken to implement the
recommendations. If the military software problem is real, it is not perceived as urgent by
most high military officers and DoD civilian officials. Our Task Force does not undertake
to prove that it is urgent; we do tell how to attack it if one wants to.

12

%3. STARS

--

Software Technology for Adaptable, Reliable Systems

The STARS program objective is to achieve by 1995 a dramatic improvement in our
ability to build reliable, cost-effective defense software by applying known new technology.
STARS seeks improvement in methods, techniques, tools, personnel practices, and business
practices.
The Task Force examined the STARS program on several occasions during the past
two years. The program is floundering. Little has been achieved during the last several
years. OSD management has recognized the problems, and some remedial steps are under
way. It is too early to tell if the~so. steps will work.
Findings
STARS as originally formulated is a very good idea.

U
v

%
0

Members of the Task Force do not expect to see dramatic near-term research discoveries. However, many incremental improvements in software engineering have been made
over the last decade, and will continue to be made. These advances could improve our
war-fighting capability if they were practiced in DoD programs now. STARS can accelerate

their application.
OSD has not provided the vital leadership needed; until recently STARS
has lacked a director with strong technical and management ability.
The program had no permanent program manager for over a year. Consequently, it
lacked leadership, guidance, and vitality. There has been no single vision of program
objectives, no coordination of spending, no monitor to ensure that components of the
program were complementary, and no assurance that the program acted in response to
the software problems of the Services. Strong top-down leadership, both technical and
administrative, is required. A new, permanent director has recently been appointed.
The program plan has been fuzzy.
The Task Force had difficulty in identifying specific goals or plans to achieve them.
The program plan does not even recognize the existence of some major software trends,
such as the personal computer revolution. STARS should enable solutions, not develop
them from first principles. It should ident~fy possible technical approaches and tools,
harness the marketplace capability to produce solutions, and ensure early application to
relmiss ion-cr itical system developments.
The STARS Program as it stands today has become focussed at a particular, welldefined part of the military software problem - custom systems, new or converted,
middle-sized to large, whether embedded or command and control. STARS addresses
folow-hroghsoftware engineering support for ADA software.
This foctui~sing is entirely commendable and beneficial; indeed, it was badly needed.
A corollary is that the problems STARS does not address have also become clear. They
include:
13

%"*personal computers, workstations, -nd little software systems built on them
""acquisition of off-the-shelf commercial software
"osupercomputer calculations (still and perhaps forever in FORTRAT"
"*old systems not worth converting
To enumerate these is not to criticize the STARS program. It never ready could address
all the problem; it only claimed to do so as long as it was fuzzy and unfocussed.
Balance between program elements has been missing.
T)evising a single software engineering environment dominates the attention of the
program. In contrast, emphasis on multiple po3sible environments, even some that are
off-the-shelf, would serve the objectives better. Most of what the STARS program proposes
to deliver is scheduled very late in it. lifetime. Early operational milestones would better
speed transfer of the technology to the DoD and civilian practitioners.
The program is organized as uncoordinated activities; many are executed
by part-time volunteers.
An independent committee explores each activity area; little communication relates
the committee actions. For example, there is insufficient integration of the activities of the
business practices area with each of the technical areas.
STARS needs better coordination with the Services, the Software Engineering Institute, AJPO, DARPA's Strategic Computing Program and the
Strategic Defense Initiative.
All these programs have interlocking interests and development programs. Links have
not been carefully established for the input of Service needs into STARS planning and the
output of STARS back to the Services.
It is difficult to get the needed funds allocated for software engineering; if
STARS is terminated a major opportunity will be lost.
An effective STARS Program is indeed needed to accelerate the application of the best
ideas from the laboratory to weapon systems development. If an effective STARS Program
does not materialize, software risks will remain high.
Salvage of STARS may not be possible, but it should be attempted. Drastic
action is required.
Recommendations
Recommendation 1: Move STARS and rebuild it.
Create a Joint Program Office to oversee the STARS program, AJPO, and the Software
Engineering Institute. This Office should be headed by a flag-rank -military officer in
order to demonstrate DoD cammitment to provide firm oversight, resources, and control
14

%over DoD software technology efforts. This Joint Program Office should report to the
Deputy Undersecretary of Defense for Research and Advanced Technology. Locate it at
the Electronic Systems Division in Bedford, Massachusetts, with the Air Force as execitive
agent.

This management organization has been an effective technique for mustering Service
cooperation on joint efforts in the past. Examples include WIS, Joint Tactical Fusion,
JTIDS, and the Joint Surveillance and Target Acquisition Radar System. OSD ret,,ns
oversight authority, and the Joint Program Office organization will ensure that benefits
accrue across all the Services.
Recommendation 2: Task th•e STARS Office, the Ada Joint Program Office,
the Software Engineering Institute, the SDI software methodology program
element, and the DARPA Strategic Computing Program to produce a one-time
joint plan to d,-nonstrate a coordinated DoD Software Technology Program.
This plan must ensure ongoing technical exchange among the five programs.
Recommendation 3: Task the new STARS Director to define a new set of
program goals together with an Implementation plan; emphasis should be on
visible, early milestones that have demonstrable results.
This plan should emphasize widespread adoption of the best that exists today. It
should provide incremental products. It should complement what the commercial sector
is doing and focus on DoD-unique requiremrents. It should be realistic.
Recommendation 4: Direct STARS to choose several real programs early in
development and augment their funding to ensure the use of existing modern
practices and tools.

15

%4. ADA
DoD defined the Ada language (see MIL-STD-1815A) to be its common, machineindependent programming language for Dr.D-wide use in mission-critical computer applications. This intent was established in 1981 by draft versions of DoD Instruction 5000.31. A
subsequent June 1983 memorandum from Dr. Richard DeLauer, Undersecretary of Defense
(R\&E), mandated the use of Ada on all new DoD mission-critical computer procurements
entering concept definition after I January 1984 or entering full-scale development after 1
July 1984. Mr. Don Hicks, Undersecretary of Defense (R\&E), reaffirmed thie mandate to
Ada in December, 1985, a, did Secretary Weinberger in November, 1986.
The Task Force discussed Ada, its compilers, and its application in military programs
Lth the three Services, inter-Service programs such as WIS, and the acting Director of
the Ada Joint Projects Office. ThL Task Force also considered fourth-generation languages
and their implication for the Ada effort.
's

Findings
Improved Software Engineering Techniques
meth-ds and techniques have dramatically advanced
overSoftware
the last engineering
decade, yet these techniques are not generally practiced in DoD.
Ada is not merely a programming language; it is a vehicle for new software practices and
methods for: specification, program structuring, development and maintenance. Without
enforced usage of such a vehicle, the radical improvements in software :ngineering will not
mov" rapidly into use. Standardization on a language is the best way to introduce the new
practices rapidly.
Ada, The Standard Language
It is a major technology step forward for the DoD to iisist that all software
be built in a "ligh-.evel language. It is a major management step forward to
stand; rdize or. a single high-level language.
It is not simple to do so; Fnr4tan and Cobol will each survive in some military
applications. The driving reasons to standardize new develcpment in one high-level
language remain valid. Spei.i•,..ally, the quality of th- resulting software will be higher.
Enhancement of function, adaptation to environmental changes, and fixing of errors will
be less buggy ax d cheaper.
Even where exceptions to the use of Ada are granted, all software can and should be
designed using Ada as a design language.

16

S

%Ada was designed by the DoD to be that standard language; It to the best
candidate for standardization available today; It promises to remain so for the
foreseeable future.

1W

Ada's constructs support modern software technology and discipline. Ada supports the
eviolution and maintenanceu~f reusable software, portable software, and real-time software.
The langiage definition is precise enough. Other candidate languages have many more
deficiencies than Ada with respect to the DoD's needs.
Ada is adrxittedly complex. This complexity has contributed significantly
to the slow maturation of the language and of Its compilers and tools.
E~nough Ada compilers now exist to demonstrate they can be built. Because of
language complexity, current compilers execute slowly in comparison to a good Fortran
compiler. However, the compilers are doing more checking, and pointing out errors to
the programmer; this is co.t-effective. Engineering refinement of compilers will yield
acceptable, even good, Ada compiler speed in the near term. Moreover, modern partialcompilation techniques today reduce the impact of raw compiler speeds.

IV

Due to Ada's complexity, the code generated by current Ada compilers Is
not yet highly optimized.
Again, engineering refinement will produce optimizing compilers in the near term.
Whereas Ada application code can be quite slow if all dynamic checking is enabled, most
checking can be turned off in the production version of application code. There is no
technical obstacle to achieving optimized code for applications written in Ada.
The DeLauer mandate to use Ada was premature; it could not be followed
In 1983 because of slowv maturation of the language and Its compilers.
Consequently it became toothless. The compilers have been developed to a point that
the mandate can be implemented now; it should be.
Switching to Ada necessitates an up-front investment In order to reap longer
term benefits.
One cost is education. Teaching Ada also implies teaching the new software engineering
practices and disciplines. This must be done anyway. Forcing this learning is a major
moiefor adopting Ada quickly and extensively.
Computer time costs will be somewhat higher because of the slower compiling. These

04

costs

are transient and will go down as Ada programming environments are widely

installed, as the software tools mature, and as hardware cost/performance continues to
drop.

II

Although Incurring the up-front costs Is wise for DoD, Individual program
managers and contractors have no Incentive to do so.

The costs of training, compiler and tool acquisition, and running the current immature

compilers are present and readily measured, whereas the benefit is future and more difficult

17

%to measure. Adoption v,.ust therefore be mandated by high management.
Ada Is being iuccessfully used today in military programs, such as AFATDS.
At least sixty-four validated compilers exist, with more in the wing. Moreover, Ada
is not just a DoD captivo language. Civilian commitment to Ada is emerging. It i
noteworthy that the majority of compilers for Ada are built with private, not DoD, funds.
One cannot predict, however, that Ada will become the standard language for civilian data
proceasing, as Cobol did. Too many different forces are at work.
Fourth-Generation Lanr .agea
Fourth-ganeration languages are application-specific program generators;
because they are not general purpose, they are not in competiton with Ada.
The term fourth-generation is a misnomer. It has been used to characterize a wide
variety of languages which are not descendants of the third-generation, general-purpose
languages. The term encompasses application-specific languages such as database languages and electronic spread sheets, program generators, non-procedural languages, and
even artificial intelligence languages such as Prolog. Each language is designed to be
applied to problems in a limited domain. Therefore the fourth-generation languages do
not compete with Ada.
If an application is well-matched to a fourth-generation language, the cost of
realizing the application can be a hundredfold less expensive than programming
it in any general purpose language, Including Ada.
Spreadsheets are routinely used to accomplish tasks in minutes that -A.uld require
hours of work in a general purpose laWrguage. Similarly, an exploratory artificial intelligence
(AI) task may be programmed in an Al language in days versus months in aiy general
purpose language.
A weapons system development is not one task in a single problem domain;
the Task Force is skeptical that any fourtb-generation language is well-suited
to such applications.
Note that some of the high-risk tasks in a weapons system may be advantageously
prototyped in a fourth-generation language to experiment with algorithms or software
structure before actual development commences. Similarly, some single-domain missioncritical applications, such as some intelligence data processing, may be cost-effectively
implemented in a fourth-generation language such as a database provides.
p

Some efforts to develop large software systems entirely in a fourth-generation language,
such as the New Jersey Motor Vehicle Registration System, have been unsuccessful.

18

%DoD Management of Ada
Only top DoD management can sustain a policy and program for incurring
the costs and risk of early DoD use of Ada.
Contractors incurring the up-front costs must have assurance that invatment in Ada
tooling will pay off. Programs must plan for long-range cost and quality improvements.
The Ada Joint Program Office (AJPO) is the DoD's focal point for policy
and coordination of Ada standardization, validation, and language control; it
has done a commendable job in achieving its technical objectives.
The AJPO has maintained a stab!e language definition. It has defined a comprehensive
validation suite of ltuiliage conformity tests. Note that language conformity does not
ensur- that, a compiler is robust, acceptably efficient at compile time, or capable of
generating correct or efficient code for real applications. Concentrated focus on language
conformity has slowed compiler maturity along these other dimensions. The AJPO has
also performed a communication function with its Ada Information Clearinghouse.
Definition of the Ada language and development of compilers has been
successful; the next step is to implement DoD applications in Ada.
This step is mainly acquisition management and is discussed elsewhere. AJPO can
best assist by providing truthful, complete, and candid information about compiler and
application activity.
The next technical step is to develop Ada support tools beyond compilers
and to integrate them with one another and with the underlying operating
system.
The unclear boundary between the AJPO and the STARS programs' charters has led
to some confusion of who should develop what support software technology.
Ada support tools
Ada ham been overpromised.
The Ada language embodies much current software technology. But to build application code that is portable and reusable requires disciplined use of the new engineering
practices and tools as well. Such support tools are not yet integrated with the compilers.
Support tools are needed for such activities as:
0 software documentation writing and formating;

* version and configuration control of both software and documentation;
* maintaining development history in a way that links requirements, design specification, code documentation, source code, compiled code, problem reports, code

19

%changes, tests, and test run results;
debugging; and
• project schedule and effort management.
As a consequence, non-technical managers of programs are expecting results
that no high-level language can by Itself deliver.
Environments that integrate such tools are not yet available for Ada. They ae likewise
available only piecemeal or not at all for Jovial, C, CMS, Fortran, etc.
Acquiring these environments Is the next step.
The Dob is supporting efforts to develop such environments. Environment design is
more difficult than language definitio.'. Efficient environments depend integrally upon the
host operating system, yet one wants tool portability and language independence. We
expect that market forces will produce a variety of environments around Ada if the DoD
maintains firm commitment to the language.
Recommendations
Recommendation 5: Commit DoD management to a serious and determined
push to Ada.
Management waffling is more likely to cause a failure in Ada than are technical or
acceptance problems.
Specifically, the DoD should
* Finalize and issue software language policy which reaffirms and details the policy set
forth in the DeLauer and Hicks memoranda, and the Weinerger speech.
* The DoD should establish Ada as the common, machine-independent, missioncritical computer system programming language for DoD-wide use. The mandate
should not be limited to embedded computers in weapon systems.
* Each DoD component should develop and implement a plan for cutting over to
full Ada usage. These plans should provide for support software, education, and
training of military, technical, and management personnel.
Stiffen practices for granting exceptions from Ada policy so that exceptions are difficult
now and become increasingly difficult with time.
* Mandate that where implementation exceptions are granted, software should nevertheless be designed using Ada as the design language.
Recommendation 6: Move the Ada Joint Program Office Into the same
organizationas STARS and the SEI.

I20

The major objective for Ada has become one of implementation -

using the Ada

%language for DoD systems - now that the AJPO has technical control of the language.
Common management of these three programs will strengthen each and permit easy
coordination of common goals and objectives.
Recommendation 7.: Keep the AJPO as the technical staff support agent
for the DoD's executive agent.
Specifically, it should:
* Continue firm control of the Ada language definition, permitting only minimal, if any,
change to the now-stable language for the next two years.
* Continue Ada compiler validation. In enlarging the validation suite, give priority to
adding tests that are representative of real applications.
• Encourage the definition of Ada working vocabularies. Promulgate them.
* Change the AJPO communication function to reduce the overselling Cf Ada and to
gather and report accurate, credible information on Ada tools and usage. It should:
* provide information on the existence and performance of Ada
environments.

)mpilers and

* documient experiences of the application of Ada including both success stories and
lessons learned.
a disseminate significant Ada information via newsletters, on-line data bases, books,
articles, workshops, conference presentations, and videotapes.
a continue to act as a clearing house recording availability of existing and reusable
Ada packages, or even entire tools, and objectively reporting on the experience
with thaG software.
* Continue the effort to establish a performance test suite as a companion to the language
conformity test suite.
"*Locate software measurement techniques and tools. Publicize them and make these
tools and techniques available to project managers using Ada.
"*Initiate significant measuring and recording of lifecycle costs for several major Ada
application programs.
" Continue to encourage the development of programmer support environments built on
Ada, but be slow to standaidize environments. Let a winner emerge first. In particular,
ALS, ALS/N, AlE and CAIS should not be standardized until and unless experience
with prototypes shows implementations to be effective and efficicnt.
Recommendation 8: DoD policy should continue to forbid subsetting of the
Ada language.
There must be only one definition of the language. Further, all compilers should

21
....
...

%correctly process the entire language; the validation process should continue. Without
this, much of the benefits of standardization will be lost. However, organizations and
educators should be encouraged to establish and publish useful working vocabularies to
simplify the task of learning and using Ada.
Specific working vocabularies may change as the technology matures. We see three
categories - acceptable vocabularies, vocabularies to be used with certain constraints,
and vocabularies that might not be used until the compilers and runtime environments
mature. Tasking exemplifies the second category. The use of tasking has to depend upon
the performance capability of the specific compilers. In the third category, today the Use
s tement might be limited because an inordinate amount of recompile time is needed.
Recommendation 9: The DoD should Increase Investment in Ada practices
education and training,for both technical and management people.
Each DoD component's implementation plan should include provisions for extensive,
in-depth Ada education and training. Do not underestimate the education and training
required for managers, analysts, and administrators, in addition to that for software
engineers.
Recommendation 10: Allow fourth-generation languages to be used where
the full life-cycle cost-effectiveness of using the language measures more than
tenfold over using a general-purposelanguage.
Marginal increases should not dictate using such languages, especially for long-lived,
production software. The estimated cost of a program element built in a fourth generation
language must include full life-cycle cost, including development, integration of the
program element with others built in Ada, as well as the increased maintenance costs
of support software written in multiple languages.

22

111, C

%5. Strategic Defense Initiative Software
Findings
The Strategic Defense Initiative (SDI) has a monumental software problem
that must be solved to attain the goals of the initiative.
It is critical quite out of proportion to its cost, lb:ause hardware has high replication
costs and software does not. Initial contractor proposals therefore largely ignored it.
The software problem has already received considerable public attention
and notoriety.
No program to address the software problem is evident.
Recommendations
Recommendation 11: Focus a critical mass of software research effort on
the software needs that are unique to the SDI objectives.
The SDI should use what STARS, the SET, DARPA and industry produces. Much of
the software problem faced by the SDI is due to the magnitude of the required software
and the complexity of controlling the interactions of so many components with very rapid
communication and response. This is not a unique requirement. The SDI should determine
what portion of their software problem is unique and concentrate its attention on solving
the SDI-unique problem, not the general software technology problems.
Recommendation 12: Use evolutionary acquisition, including simulation
and prototyping, as discussed elsewhere in this report, to reduce risk.

23

%6. DoD and the Civilian Software Market
Findings
The civilian market for software Is today substantially larger than the size of
the DoD market, although the DoD continues to be the largest single customer
for computer software [Jorstad, 1984; Boehm, 1986].
This new phenomenon requires a radical update in DoD thinking, policies,
and procedures.
We find that in policy drafting and debate, the mass civilian market is generally
ignored.
One important implication is that DoD cannot, as it did with Cobol, create a de facto
standard and impose it on the civilian market. This is not to say that the civilian market
will not adopt tools and methods from DoD where they are perceived as advances. We
merely observe that it will not adopt them Just because DoD has.
A second implication is that DoD, although it will necessarily lead in some aspects of
the technology such as interfaces to sensors and effectors, cannot expect to lead in most
aspects of software technology development.
A corollary is that wherever DoD's software methodology diverges from what the
civilian market is evolving by competitive selection, it will have to support its own
idiosyncratic methodology all by itself, without the resource commitment of the larger
economy.
Computer Security and Commercial-Off-The-Shelf-Software. Computer security
requirements are frequently cited as a reason why commercial off-the-shelf software cainot
be used. The National Computer Security Center in NSA has published criteria for
assessing the effectiveness of security controls in ADP systems (DoD CSC-STD-001-83).
The Center is also working on guidelines for matching the appropriate level of security
controls to a problem. With support from the National Computer Security Center,
industry is now developing operating systems, trusted computer programs, guards, and
other software and hardware products with security controls built in to the hardware and
software and is seeking certification for these products. This approach will provide more
standard and cheaper ways of dealing with computer security than the current practices
of custom-tailoring systems. It will also facilitate integration of computer security controls
,with communications security systems through the use of low-cost encryption devices and
standard interfaces for network control.
Enlisting Industry and Universities. Our Terms of Reference explicitly charged us
with suggesting how DoD can enlist the efforts of universities and industry in its software
technology advance. Our response must be that this charge itself assumes 'he bygone
situation. It is now the case that DoD, in mapping its software thrust, must in part assess
which way the technical thrust of the larger community is going, and diverge only when
24

04V

%absolutely necessary.
On those aspects where DoD is truly innovating technically, it can enlist the larger
community by the very excellence of the innovations - the competitive marketplace is
responsive to innovation.
Recommendations
Pecommendation 13: The Undersecretary of Defense (Acquisition) should
adopt a four-category classification as the basis for acquisition policy
We see vast differences in the software systems that DoD buys and builds. We
recommend that these differences should be explicitly recognized by an official classification
into four major classes according to uniqueness and novelty. Acquisition guidance, pr l"cies,
and procedures should be framed separately for each class. The classes are:
Standard:

Off-the-shelf, commercially available

Extended:

Extensions of current systems, both DoD and commercial

Embedded:

Functionally unique and embedded in larger systems

Advanced:

Advanced and exploratory systems.

Each Program Manager would classify his system, its subsystems, major components,
major increments, and phases into one of these classes, with the burden of proof being to
show why the element should be in a higher class instead of a lower one.
Table 5.1 provides attributes of the four categories. Table 5.2 illustrates the categories
by classifying some current acquisitions.
It may also be wise to establish categories by size (lines of source code) for uniformity
in description. A possibility might be:
Modest:

Under 2000 LOSC

Small:

2000-10,000 LOSC

Medium:

10-100 KLOSC

Large:

Over

100

KLOSC

Then a planned undertaking could be characterized for policy or procedure purposes
as, for example, a "Medium Extended-Class software project."

%Recommendation 14: The Undersecretary of Defense (Acquisition) should
develop acquisition policy, procedures, and guidance for each category.
Recommendation 15: The Undersecretary of Defense (Acquisition) ard the
Assistant Secretary of Defense (Comptroller) should direct Program Man.agers
to assume that system software requirements can be met with off-tk e-shelf
subsystems and components until it is proved that they are unique.
The cheapest way to get software is to buy it in the commercial marketplace rather
than to build it.
The fastest way to get software is to buy it in the commercial marketplace rather than
to o uild it.
The surest way to get robust, maintained, supported software is to buy it in the
commercial marketplace rather than to build it.
Even though commercial software is often delinquent in these latter respects, competitive pressures get the delinquencies fixed. Custom-built software has historically been
notoriously bad in these respects, without the same pressures to fix it.
Hence the DoD-wide assumption should be that a commercial product will be usable if
the function is similar to that required - perhaps with modifications or extensions, perhaps
with extra documentation, perhaps with different support and maintena,.nce arrangements.
The first investigation when requirements begin to be formulated should be into the market,
to ,-Le what is available already. Even if the best off-the-shelf product is not ultimately
used, adopting it for pilot use will help radically in setting specificrLtions for the custom
product.
Recommendation 16: All the methodological efforts. especially STARS,
auld look to see how commercially available software -ools can be selected
I standardized for DoD needs.
Although end-user systems, especially embedded ones, will often need to be specialized,
and : 'rhaps custom-built, it will rarely be justified for DoD to custom-build the tools with
which its software is built. The assumption should be that marketplace tools will be used.

rM

01
28

%7. DoD In a Sellers' Market for Software
Findings
Because of the explosion of the commercial software market, DoD now Is
In a sellers' market for software-building.
Just as the DoD need for software is growing exponentially, and its software-skilled personnel grow more slowly, the same is true of the commercial software n-arket. Programmers
are in short supply, programming managers even scarcer, and software system designers
very scarce. Hence the companies that have these skills, and have them organized into
functioning, equipped teams, have many choices as to how best to market their services.
DoD is perceived as a poor customer, and the stable of DiD custom software
vendors stays small even though the requirement grows radically.
Poor Return. Building custom software for DoD has a poor profit margin. In calculating
proper profit levels for cost-plus-incentive contracts, DoD tends to use the same marginm
for software development as for hardware development, although the latter is customarily
followed by a production cycle at acceptable total profit levels. Ten percent profit on sales
is considered high in DoD, whereas it is grossly unacceptable in computer industry pricing
on software.
Weak Incentives for Productivity.
Even on fixed-price software contracts, there
are only weak incentives to manage for higher productivity or to invest company money
in capital tooling that will save labor cost. High productivity and high quality are not
rewarded by DoD except at the time of contractor selection. If "excessive" profits result
from high productivity on firm-fixed-price contracts, the profits are readjusted after audit.
The standard for "excess" is the same as that for hardware development, despite the
absence of the production cycle.
Heavy Regulation. DoD Directive 5000.29, "Management of Computer Resources in
Major Defense Systems", DoD STD 2167, "Defense System Software", and the proposed
new Federal Acquisition Regulation FAR 27.4, "Data Rights", and the proposed DoD FAR
Supplement 27.4 have as their purpose to ensure fair, consistent, and open competition,
and to get the most capability for each dollar spent. In practice, however, they inhibit the
use of standard commercial practices in software acquisition and maintenance. They also
encourage the building of new software rather than purchase of off-the-shelf items.
Unreasonable Rights-In-Data Requirements.
DoD FAR Supplement 27.4, as
proposed for public comment by 10 January 1986, set forth demands for DoD ownership
of rights to programs, documentation, tools, and methods that were
"*formulated only in terms of hardware, entirely ignoring the different circumstances
of software,
"*completely at variance with commercial practice for software,
29

%""probably

illegal under U.S. Copyright law, and

"*destructive of most vendor incentive to hivest in better tools and methods.
We were convinced that the misfit between the proposed supplement and software is
entirely unintentional, and so the Task Force made timely presentations to the Undersecretary of Defense (Acquisition) and to the Assistant Secretary for Acquisition and Logistics
during the comment period. If the proposed supplement were to be adopted, however,
it would be another powerful disincentive for vendors to bid on DoD custom software
development.
Constant, Small, Stable of Vendors. Given the regulatory and financial structure
of DoD contracting outlined above, a vendor can participate in substantial DoD software
contracting only if it:
* has staying power to average over crests and troughs in contracting business,
* has a management superstructure to cope with the regulatory overhead, including
a Program Control Management System, and specialized accounting to meet DoD
STD 7000.2,
,S.

* has an infrastructure of technical specialists to deal with configuration management
per regulation, documentation per regulation, acceptance testing, etc.

2

%therefore has a critical mass of skilled people sc. -hat it can carry several contracts
a' once, both to smooth crests and troughs and to pay for the administrative and
technical superstructure necessary to cope w'h the regulatory overhead.
As a result, we estimate there to be only about two dozen houses that are regularly
available to participate in the development of substantial mission-critical software systems,
and this number grows slowly. On the contrary, several vendors are seriously considering
leaving the DoD business entirely, and refusing to bid in the future.
The net result of this small stable of vendors in a time of exponential growth of work
is sure to be higher bids and longer schedules. It is in DoD's interests, and the nation's
interests, for DoD to make itself an attractive customer. We believe the net costs to the
nation of weapon-system software will be lower if it does. Present practices are penny-wise
and pound-foolish in many petty ways.
Recommendations
Recommendation 17: DoD should devise increased productivity incentives
for custom-built software contracts, and make such incentivized contracts the
standard practice.
A new contracting form, part-way between fixed-price and cost..plus-fee, should be
devised. For instance, on a cost-type contract, a pioductivity figure is usually bid.
Competition in vendor selection keeps the figure from being unreasonably low. So
reimbursement might be structured to split any savings due to increased productivity

30
01

%evenly between the buyer and the seller.
Another new contracting form that we recommend DoD consider would be to guarantee
a quantity buy of some software product and to request bids solely on a per-copy price. Here
the vendor would bear all the non-recurring costs and recover gains on capital investment
and productivity enhancement. Ada compilers and softwa re development and maintenance
environments are examples that could be purchased or licensed this way.
Recommendation 18: DoD should devise increased profit incentives on
software quality.
One such incentive could be a sliding profit margin based on the quality of the delivered
complete software product. This requires quality metrics, recommended below.
Another kind of incentive could be introduced by requesting contractors to bid a percopy-per-year fixed license fee including maintenance. In this way, high quality resulting in
low maintenance would provide financial rewards to the contractor and operational rewards
to the users.
Recommendation 19: DoD should develop metrics and measuring techniques for software quality and completeness, and Incorporate these routinely
in contracts.
There are today no metrics for source-code quality, object-code quality, documentation
quality, etc. Part of the STARS methodological effort should be addressed to such metrics,
for Ada programs in particular.
Meanwhile, there are techniques for judging the over-all quality of complex performances outside the computer field. There is wide agreement as to what quality is, and
skilled practitiol.,ers make similar judgements when presented with products to judge. Even
today software quality can be judged by panels of trained judges, just as such panels judge
Olympic diving, skating, and acrobatic performances. DoD should immediately begin
testing such panel methods for consistency and reliability, and, if they work, begin using
such judgements for quality incentives.
DoD buys software for operational systems that will be used ior a decade or more.
The software must be maintainable and changeable. DoD is willing to pay the price for
complete software products, but all too often it accepts less because of schedule slippages
and operational needs.
Complete software products should be mandated in contracts to include:
* specifications thae describe the actual software structure as built, as opposed to
that originally specified,
• documentation showing the structure and organization of the software,
* source code that is properly structured, well modularized, and well ccman-ented,
especially in procedure headers and variable declarations,

31

%ccross-reference documentation that traces articles in the specifications to the
corresponding source code, and vice-versa.
Recommendation 20: DoD should develop metrics to measure Implementation progress.
Such metrics would help ensure that costs and schedules are being met and that
complete products will be delivered. They might include, for example, program size, software complexity metrics, personnel experience, testing progress, and incremental-release
content. Development of such has in the past been part of the STARS plan; it should
continue to be. Meanwhile, panel-judging techniques as discussed above can be applied to
progress as well as to quality.
Recommendation 21: DoD should examine and revise regulations to approach modern commercial practice insofar as practicable and appropriate.
Recommendation 22: DoD should follow the concepts of the proposed FAR
2T.4 for data rights for militariy software, rather than those of the proposed
DoD Supplement 27.4, or it should adopt a new "Rights In Software' Clause
as recommended by Samuelson, Deasy, and Martin in Appendix A6.
The legal problem is highly technical. Two good solutions, the arguments for proposed
FAR 27.4, the concerns about t•he 'clarity fad applicability of the proposed DoD Supplement
27.4, and the arguments for a new clause L all skillfully set forth in the SEI Technical
Report incorporated herein ,ts Appendix 6A.
To those technical arguments we would add an economic one: the proposed DoD
Supplement 27.4 is, ;ntentionally or otherwise, grabby in spirit and effect. No fair-minded
person would considt!r it to propose equitable treatment among vendors, or between DoD
and vendors. Its lack of clarity and clumsiness of drafting will occasion litigation. The net
effect will be to further shrink the vendor pool, at gre%t cost to DoD and the taxpayer.

32

%8. A New Life-Cycle Model for Custom DoD Software
Findings

I

The most common present method of formulating specifications - issuing
a Request for Proposal, accepting bids, and then letting a contract for software
delivery - is not in keeping with good modern practice and accounts for much
of the mismatch between user needs and delivered function, cost, and schedule.
As discussed above under Current Trends, we now understand the importance of
iterative de ,elopment of requirements, the testing of requirements against real users' needs
by rapid prototyping, and the construction of systems by incremental development, with
early incremental releases subjected to operational use.

The Task Force finds that Directive 5000.29 and STD 2167, as interpreted,
have made it difficult to apply these modern methods.
Although some parts ef the recent Draft DOD-STD-2167A appear to encourage modem
methods, the draft as a whole continues to reinforce exactly the document-driven, specifythen-build approach that we believe causes so many of DoD's software problems.

Recommendations
Recommendation 23: The Uv~dersecretary of Defense (Acquisition) should
update DoD Directive 5000.29, "Management of Computer Resources In Major
Defense Systems", so that it mandates the Iterative setting of specifications,
the rapid prototyping of specified systpems, and Incremental development.
We propose that the iterative development, of specifications can be reconciled with
the needs of fair and open competition by letting two level-of-efforts contracts for the
specifying and prototyping of major software systeLms. After prototyping is complete and
specifications formulated, a software production contract can be put out for bidding in the
usual process.
An alternative for the specification process is to let a separate contract to a government
contractor for specification, with the specifying contractor excluded from bidding on the
build.
The iterative development of specifications is a small part of the total cost of a major
software system, usually less than 10%, but it has profound effectE on the procurement
cost, life-cycle cost, schedule, and function of the product.
We believe the procurement cycle should be modified so that the requirements remain
unfrozen and subject to alteration until the cost and performance effects of their provisions
'can be known from early product design. This means final requirements would not be
frozen until perhaps one-third of the way through the procurement period, a substantial
departure from present practice.

Recommendation 24: DoD STD 2167 should be ,'urther revised to remove
Pi3

33

%any remaining dependence upon the assumptions of the Owaterfall* model and
to Institutionalise rapid prototyping and Jucremental development.
Recommendation 25: Directive 5000.29 and STD 2168 should be revised
or superseded by policy to mandate risk management techniques In software
acquisition, as recommended In the 1983 USAF/SAB Study.
The Air Force Scientific Advisory Board in 1983 identified software risk factors and
recommended risk management techniques [Munson, 1983). While parts of the recommendations are Air-Force-specific, the ideas ae applicable to all Services. An example of how
týse risk management techniques have been incorporated into program management is
ij ,ided in Table 7.1 below. The risk-management approach provides an effective way
foz a project to determine when, where, and how much to use prototyping and similar
risk-reduction techniques.
Table 7.1
Software Risk Management Plan
1. Identify the project's top 10 risk items.
2. Present a plan for resolving each risk item.
3. Update list of top risk items, plan, and results monthly.
4. Highlight risk-item status in monthly project reviews.
5. Initiate appropriate corrective actions.
Recommendation 26: Each Service should provide its software Product
Development Division with the ability to do rapid prototyping In conjunction
with users.
The DoD software system acquisition agents are the service produc. development
divisions. Each of these divisions needs facilities and equipments to mock-up, simulate, and
build critical prototypes of the new systems being acquired. Some of the system interfaces
can be tested prior to delivery to the users. Such facilities can also serve as a place for
the developer and user to meet and refine requirements and procedures. The placement
of such facilities at the product development division level will allow their use by multiple
Program Managers who all report to the same local commander. The product development
divisions are organized along mission categories, and programs in each division will tend
to need equipment and software with similar power and capabilities. The facilities could
also be used for Ada training.
Recommendation 27: Each Service should provide its software Using Commands with facilities to do comprehensive operational testing and life-cycle
evaluation of extensions and changes.
The user commands are responsible for defining the original requirements. However,
many of the systems that are being developed are doing jobs that have never been done

34

%before. These types of systems tend to be technology-driven and must be placed in the
hands of the user as early as possible to develop new operational procedures. In the
early phases of system acquisition, a facility at the user command is needed for testing,
evaluation, training, and procedure development. Throughout the life of the system, the
user commands need the facility for testing and evaluation of changes and upgrades to
systems.
Recommendation 28: The Undersecretary of Defense (Acquisition) and the

Assistant Secretary of Defense (Comptroller) should by directive spell out the
role of Using Commands in the evolutionary and incremental development of

software systems.
The relationships between Developing Commands and Using Commands for the
different kinds of systems should be spelled out in policy statements. The role and
responsibilities of the user commands can vary with the system acquisition procedures
and the kind of system being acquired. In conventional acquisitions such as weapons,
platforms, or sensor systems, a system is developed, tested, and turned over to the user.
For command and intelligence systems, much of the development and testing can take
place at the user command. The experience of the users with the first capability built can
be used (or required) as feedback to the second increment of the system. User involvement
is obviously heavier when evolutionacy acquisition procedures are used.

35

%9. Module Reuse in DoD Custom Software
*

Findings
*

Software technology now enables the extensive reuse, even In missioncritical embedded systems, of software modules written for other systems.
The typical module size is on the order of one to two pages of source code, 25-50
source lines. Ada in particular allows modules to be used easily and safely, because the
module interface is packaged and specified separately from the body, which the user does
not ordinarily need to inspect or alter.
Module reuse requires new forms of contractor incentives, both to make
modules available for others to use, and to use them themselves instead of
building anew. Making a module reusable requires a modest extra effort in design and
development. There has to be incentive and compensation for this effort.
Module reuse requires the establishment of clearinghouses or markets where
modules can be exchanged.
Module exchange requires the establishment of standards of description of
function and of degree of testing.
Recommendations
Recommendation 29: The Undersecretaryof Defense (Acquisition) should
develop economic incentives, to be Incorporated into standard contracts, to
allow contractors to profit from offering modules for reuse, even though built
with DoD funds.
Recommendation 30: The Undersecretaryof Defense (Acquisition) should
develop economic incentives, to be incorporated Into all cost-plus standard
contrac~ts, to encourage contractors to buy modules and use them rather than
building new ones.

I;

Acquisition contracts should be structured so that contractors will be motivated to
build and sell reusable software, and to buy it. Reusable software will be successful when
contractors decide it is in their competitive self-interest to reuse software rather than to

develop it each time. The proper incentives with respect to data rights, warranties, licenses,
liabilities, and maintenance must be included in the RFPs and the contracts.
Recommendation 31: The Undersecretary of Defense (Ac quisition) and
Assistant Secretary of Defense (Comptroller) should direct ProgramManagers
to Identify In their programs those subsystems, components, and perhaps even
modules, that may be expected to be acquiredrather than built; and to reward
such acquisition In the RFP's.
Recommendation 32: The Software Engineering Institute should establish
a prototype module market, focussed originally on Ada modules and tools for
36

%Ada, with the objective of spinning it off when commercially viable.
A scheme for how such a marketplace might work, including some possible financial
and licensing arrangements, is proposed in Appendix A7.
The White Sands Missile Range is operating today an Ada Software Repository,
apparently using volunteer labor and spare computer capacity. We believe that a more
regular and reliable service must be based upon licensing and license fees, and our proposal
includes that. Rudimentary validation of compilability by the marketer may also be
necessary.
Recommendation 33: The Software Engineering Institute, in consultation
with the Ada Joint Program Office, should establish standards of description
lor Ada modules to be offered through the Software Module Market.

37
U'

-

%10. Software-Skilled People
DoD's demand for software capability grows exponentially. It does so at a greater rate
than the combined growth in the size and productivity of the pool of software personnel.
Previous DoD studies have identified personnel issues as critical elements of DoD's
software problems. These studies have made excellent recommendations on personnel
issues, but the recommendations have not been acted upon.
The Task Force recommends a new approach to the software personnel problem.
Findings
Previous Studies Have Made Good Recommendations
A number of previous studies of the DoD software problem have identified the scarcity
of in-house DoD software personnel as a critical problem. They have developed similar
sets of recommended actions for dealing with the problem, e.g.:
9 Determine DoD's needs for the various software-related skills,
* Create and maintain a skills inventory for DoD personnel,
* Create and implement more attractive career paths for DoD software personnel,
* Establish educational programs to support these career paths.
9 Analyze the factors influencing the development and retention of DoD personnel
with the appropriate mix of software-related skills,
Previous Recommendations Have Not Been Acted Upon
If these actions were vigorously pursued, they would go a long way toward solving the
problem.
However, for various reasons, DoD and the Services have not acted on these previous
recommendations. It is therefore unlikely that any effective action would result from yet
another restatement of these recommendations by this Task Force.
Wesize
believe
the pool
years.of DoD software personnel has remained about the
same
for many
The national pool of software personnel Is growing rapidly.
The number of Bachelor's and Master's Degrees in Computer Science, Mathematics,
and Statistics are shown in Table 9.1. Historically, the supply of computer science
graduates has been augmented primarily from graduates in Mathematics. These sources
of new personnel meet requirements estimated at 54,000 new graduates per year in 1983
[Hamblen, 1984] to sustain a pool of computer specialists whoee size was estimated at
299,000 in 1082 [Vetter, 1985] and whose annual growth rate is estimated at about 5%

[NSF, 1984].
38

%It appears that DoD is not competing effectively with the private sector in
attracting and retaining software personnel.
Table 9.1
Degrees In Math/Statistics and Computer Science, 1970-82
Bachelor's Degrees
Year

Math/Statistics

1970
1971
1972
1973
1974
1975
1976
1977
1978
1979
1980
1981
1982

27,442
24,801
23,713
23,067
21,635
18,181
15,984
14,196
12,569
12,115
11,473
11,078
11,599

Master's Degrees

Computer Science
1,544
1,624
3,402
4,305
4,757
5,039
5,664
6,407
7,224
8,769
11,213
15,121
20,267

Math/Statistics
5,636
5,121
5,198
5,028
4,834
4,327
3,857
3,695
3,373
3,578
2,868
2,567
2,727

Computer Science
1,459
1,588
1,977
2,113
2,276
2,299
2,603
2,798
3,038
3,055
3,647
4,218
4,935

Source: Table 12 of [Vetter, 19851
DoD needs software talent primarily to support the acquisition process
We agree with previous studies that the software personnel shortage hurts DoD most
in the area of software acquisition management.
MITRE and TRW experience have found software acquisition has been most effective
when DoD had an acquisition management cadre whose size is roughly 10% (5-15%) of the
size of the developer's staff; a cadre with a thorough understanding of software technology
and acquisition management.
The people in such a cadre are not just watchers. They add considerable value to
the software product by developing specifications, operational concept documents, and life
cycle plans; managing competitive concept definition, preparing precise RFP packages,
performing thorough source selections, including pre-award audits and independent cost
estimates; exercising prototypes for realistic user feedback; improving specifications, plans
and manuals; monitoring the effective performance of quality assurance, configuration
• management; and subcontracting and financial management, representing user interests
on change control boards.
DoD does not have adequate career paths for software professionals

39

%Some Services have no career paths; some Services alternate computer assignments with
totally unrelated non-computer assignments, thereby diffusing the officer's experience. In
many cases, software expertise is encoded as a subspecialty inflection rather than in a
primary specialty code.
Software engineering methods and techniques are advancing dramatically. It is critical
for software professionals to master these advanced methods and techniques and to keep
learning new techniques as they are developed. This is as important for acquisition people
as it is for production people. (Building architects have to know the technologies better
than most contractor people.) In-house software skills do not match those of top contractor
p 'IS.
Current deployment of the software talent pool is ineffective
Currently, many software-qualified personnel are assigned to jobs that could effectively be assigned to contractors. Many DoD software acquisitions are either. in-house
development efforts staffed entirely with DoD personnel or contracted acquisitions with
DoD staffing levels far below the needed 5-15%.
Recommendations
'VI

4

Recommendation 34: Do not believe DoD can solve its skilled personnel
shortage; plan how best to live with it, and how to ameliorate it.
The software personnel shortage will not disappear by direct DoD action. All DoD
plans should be based on the assumption that an acute skill shortage will persist.
Organization structures should be tuned, assignment policies should be adjusted, and
educational programs should be revised to produce the military and civilian cadre needed
to acquire and maintain highly complex systems.
Further, DoD should facilitate supplementing the software acquisition management
process with contractor support where the supply of in-house personnel is insufficient.

Recommendation 35: Ulse DoD people for acquisition instead of construction.
*

Instead of hoping that enough personnel can be hired and retained to satisfy the
needs of the current strategy for using software personnel, change that strategy to train
and assign available personnel to the highly-leveraged tasks, namely software acquisition
management. DoD should sharply reduce in-house software construction, extension, and
maintenance, limiting such to critical functions at operational bases, adaptation of existing

Recomendtion36:Establish mechanisms for tracking personnel skills
and rojctig prsonelneeds.
No meaningful studies have been found that catalog seasoned personnel, and no studies
havebee
fond hat nclde othuniformed personnel and government civilians.
EachSericeneeds to have, and all DoD needs access to, a database that covers its
40
........ IIJII4ý
.......

%officers, its senior and technically skilled enlisted people, and its technically skilled civilian
employees. This should show not only career history and assignments, but technical skills,
experiences, and trainings by quite fine subject codes.
From such a database each Service should not only draw for particular assignments,
but also project biennially the trends by skill, by seniority, by median age within skill, and
by years of particular skill experience. Such trends can then be assessed against projected
needs.
Recommendation 37: Structure some officer careers to build a cadre
of technical managers with deep technical mastery and broad operational
overview.
Where possible, operational assignments should be chosen to give intense systemusing experience in real operations, development/acquisition assignments should be L..n
related software systems; and education assignments should focus on new technical and
management approaches.
Recommendation 38: Enhance education for software personnel.
DoD should implement the education and training necessary for its software acquisition
management personnel to master both software technology and acquisition management.

41

%11. Appendices

Al. The Task Force

A2. Terms of Reference

A3. Meetings and Briefings

A4. Documents Studied

AS. Software

-

Why Is It Hard?

A6. Proposal for a New "Rights in Software' Clause

A7. Proposal for t Module Market

42

-- -- - -

%Appeu,, ix Al
DEFENSE SCIENCE BOARD SOFTWARE TASK FORCE
Members and Key Persons
Dr. Donald Hicks, USD(R\&E), Sponsor 202-695-6639
Dr. James P. Wade Jr., Deputy USD(R\&E), Acting ASD(A\&L), Sponsor
Dr. Victor Basili, Member
Chairman, Department of Compute: Science
University of Maryland
College Park, Maryland 20742
301-454-2002
Dr. Barry Boehm, Member
TRW Defense Systems Group
One Space Park, R2-2086
Redondo Beach, CA 90278
213-535-2184
Ms. Elaine Bond, Member
Senior Vice President
Chase Manhattan
One New York Plaza, 21st Floor
New York, NY 10081
212-676-2982
Dr. Frederick P. Brooks, Jr., Task Force Chairman
Kenan Professor of Computer Science
University of North Carolina at Chapel Hill
New West Hall 035A
Chapel Hill, NC 27514
919-962-2148
Mr. Neil Eastman, Member
IBM Federal Systems Division
18100 Frederick Pike
Building 929, Room 1C12
Gaithersburg, MD 20879
301-240-2170
MGen. Don L. Evans, Member
USAF (ret)
President
Tartan Laboratories
461 Melwood Avenue
Pittsburgh, PA 15213
412-621-2210

43

%Dr. Anita K. "ones, Member
Tartan Laboratories
461 Melwood Avenue
Pittsburgh, PA 15213
412-621-2210
Dr. Mary Shaw, Member
Software Engineering Institute
Carnegie-Mellong University
Pittsburgh, PA 15213
412-578-7731
Mr. Charles Zraket, Member
President
The MITRE Corporation
Burlington Road
Bedford, MA 01730
617-271-2356
Dr. Edward Lieblein, Government Representative, OSD
202-694-0208
LTC Bill Freestone, Government Representative, Army
202-694-7298
Mr. Marshall Potter, Government Representative, Navy
202-697-9346
LTC Dave Hamond, Government Representative, Air Force
202-697-3040
Major Susan Swift, USAF, Executive Secretary
202-695-7181
CDR Chris Current, DSB Secretariat Representative
202-695-4157
CDR Mike Kaczmarek, DSB Secretariat Representative
202-695-4157
Mr. Robert L. Patrick, Contractor Support
Willow Springs Road
Star Route 1, Box 269
Rosamond, CA 93560
805-256-4444
Mr. John K. Summers, Contractor Support
The MITRE Corporation W90
7525 Colshlre Drive
McLean, VA 22102
703-883-6146

44

%RESEARCH ANO
*

ENGINEERING

CDSB)
2NOV W

MEMORANDUM FOR CHAIRMAN, DEFENSE SCIENCE S!C)RD
SUBJECT:

I

Defense Science Board Task Force on Software

You are requested to form a Task Force on Software.
Software costs are projected to increase substantially in the
next decade, and the cost of software development is becoming an
increasing fraction of total development costs of many types of
weapons systems. In addition, the testing of software to prove
performance is becoming increasingly difficult and-time
consuming, leading to delays in aystem deployment.' The need for
software productivity improvement is well recognized.

The task Force should address this broad problem and

identify and assess the following:
A.
Assess and unify the conclusions and recommendations
of the various recent studies of military software problems.
B.
Examine and discuss the theoretical and practical
reasons that make software costs high including the design and
analysis of tests.
C.
The probable effectiveness of the proposed DoD STARS
program at addressing military software problems, and the
relative priority of the components of the STARS program
(suggest alternative to STARS if necessary)
D. Ways of enlisting industry, Service laboratories, and
university efforts in-programs aimed at software productivity.
E.

The probable effectiveness of the STARS program and

U.S
inerntioalcompetitiveness in software production.
P.
Hw touselimited R\&D funds to make the biggest
impoveentin

hedevelopment of military software

capabiities

%2
G.

Implementation concepts for an incremental,

evolutionary approach in case an all-out assault on the software
problem cannot be funded.
Based on the above assessments, the Task Force should make
specific recommendations for significant improvements in the way
the DoD manages and develops software.
I would appreciate a
report in approximately.,ýx months.

Dr. James P. Wade, Jr.., Assistant Secretary of Defense
(D\&s), is sponsoring this study. Dr. Frederick P. Brooks has
agreed to serve as Chairman.
Major Susan Swift, USAF, will
serve as Executive Secretary.
Commander Chris Current, USN,
will be the DSB Secretariat representative.
It is not
anticipated that your. inquiry will go into any "particular
matters" within the meaning of Section 208 of Title 19,
U.S. Code.

,

Enclosure
Proposed Membership

46

4

%I

Appenadix AS
~

Condensaed Briefings anad Mimnutes

-

The complete mnutee are on file Inthe DSB Office. The meetings arn abstracted below.

Date

Place

Subject

arlefo,

Orgaiasationt

11 March '856

Pentagon

Confilct of Interest
Army Study
Al' Study
STARS Program

David Ream
LTC SistI
John Munson
Joseph Bats

DoD Counsel

22 April '.65

Pentagon

Planning of Study

28429 May '8s

ýMTRE, McLean

Navy Software Perspective
Tactical Flag Corn. Ctr.
Dicuaslon re Task Fares

USA
OSDSTARS

COMO Harry Quest

USN

CMDR DeMaree

USN TFCC

Dr. James P. Wade, Jr.

OUSDR\&E

Assessment of STARS Conf.
SEX/STARS Relationship
1982 DoD Joint SW Study
Al' Software Perspective

BG Alm\&Saliabury
Joseph Bats
Dr. Edward Lieblela
Dr. Barry Boehm
Dr. Mary Shaw
Larry Druide
BG Dennis Brown

USA Sys Command
STARS
OSD

WIS Program

John Glilligan

Executive Session

Gene Frank
Don ONeil
Don Latham
Dr. James Wade

Deputy SPO Director
GTE
IBM
ASD(C31)
Acting USDR\&E

Report Discussion

Dr. Danny Cohen

ISI

Stategic Computing Initiative
Strategic Defense Initiative
Compet. In Contracting Act
STARS

Dr. Steve Squires
Ma4. David Audley
Wayne Wittig
Dr. Jack Kramer

DARPA
SDI
OASD(A\&L)
OSD-STARS

Pentagon

DFARS 27.4
SW Test \& Eval. Project

M1. Pamela Samnuelson
Dr. Rich De~io

SEX
Georgia Tech

22 January '86

Pentagon

Report Discussion

Dr. Hicks, Task Force

USDR\&E

4-5 March '86

MITRE, McLean

Discussion of DSB Briefing

15 April '86

MITRE, Bedford

Report Discuosson

27 May '86

SEX

Report Discussion

Army Software Perspective
STARS Program

24 June '85

Pentagon

8 July -85

NUTRE, McLean

22-23 October '85 MITRE, McLean

26 Novemb,,.

'85

47

SEX
Rational Technology
USA.F

%Appendix A4 -

Documents Studied and References

Bailey, Elizabeth, et at

An Assessment of the STARS Program, SeptemberOctober 1985 (IDA Memorandum Report M-137)

Barbacci, M.R. et \&I

"Tho Software Engineering Institute: Bridging Practice
and Potential,* IEEE Software, November, 1985

Boehm, Barry W.

"Understanding and Controlling Software Costs,*
Informetion Procwtisn 86, H.J. Kugler, ed., Amsterdam:
Elsevier Science Publishers B.V. (North Holland)

Brocks, Frederick P.

"No Silver Bullet,* Information Prcssc•. 86,
H.J. Kugler, ed., Amsterdam: Elsevier Science Publishers B.V.
(North Holland). Reprinted in Computer, April, 1987

DeMillo, Richard A., et at

Software Engineering Environments for Mission-Critical
Applications - STARS Alternative Programmatic
Approaches
(IDA Paper P-1780) August 1984

DoD

DoD Directive 5000.29 Management of Computer Resources in
Mijor Defense Systems, 26 April 1976
Draft DoD Directive 5000.29, 15 January 1986
DoD-STD-2167 Defense System Software Development, 4 June 1985
Draft DoD-STD-2167A Defense System Software Development,
1 April 1987
DoD Directive 3405.1 Computer Programming Language Policy,
2 April 1987
DoD Directive 3405.2 Use of Ada in Weapon Systems 30 March 1987
Strategy for a DoD Software Initiative, 1 October 1982
Software Technology for Adaptable Reliable Systems (STARS)
Program Strategy, 1 April 1983
STARS Implementation Approach, 15 March 1983
48 CFR Parts 214, 215, 227, and 252 Revised Defense Federal
Acquisition Regulation Supplement "Technical Datae
10 September 1985

Druffel, Larry E., et al

Report of the DoD Joint Service Task Force on Software

48

%Problems, 30 July 1982
Frewin, Gillian, et \&I

"Quality Measurement and Modelling - State of the Art
Report,* REQUEST Consortium, Esprit Project ESP/800 Strasbourg,
8 July 1985

Hamblen, John W.

Computer Manpower - Supply and Demand by States.
Quad Data Corp., Tallah~ssee, 1984

Jones, Victor E., et al

Final Report of the Software Acquisition and Development
Working Group, July 1980

Jorstad, Norman D., et al

Report of the Rights in Data Technical Working Group
(RDTWG) 23 January 1984 (IDA Record Document D-52)

Lieblein, Edward

"The Department of Defense Software Initiative - A Status
Report," Communications of the ACM,; 29, 8, August, 1986

Munson, John B., et al

Report of the USAF Scientific Advisory Board Ad Hoc
Committee on the High Cost and Risk of Mission-Critical
Software December 1983

NSF

"Projected Response of the Science, Engineering, and
Technical Labor Market to Defense and Nondefence Needs:
1982-87." NSF Report NSF84-304. 1984

Parnas, David L.

"Designing Software for Ease of Extension and
Contraction, IEEE Trane on SE, 5, 2, March 1979, 128-138

Redwine, Samuel T., et al

DoD Related Software Technology Requirements Practices,
and Prospects for the Future
(IDA Paper P-1788) June 1984

Samuelson, Pamela

Toward a Reform of the Defense Department Software
Acquisition Policy (Working Paper)

Selin, Ivan, et at

Interim Report on Air Force Base Level Automation
Environment, National Research Council, National Academy
Press, June 1985

Taft, William H., IV

Memorandum to the Joint Logistics Commanders,
12 August 1985

Vetter, Betty M.

The Technological Marketplace: Supply and Demand
for Scientists and Engineers. Scientific Manpower Commission,
Washington, 1985

Vick, Charles R., et at

"Methods for Improving Software Quality and Life Cycle
Cost," AF Studies Board, National Academy Press, 1986

49

%Weinberger, Casper W.

"Remarks Delivered at the Ada Expo 1986"

Yaru, Nicholas, et at

Army Science Boari Study on Acquiring Army Software, 1983

Zracket, Charles A., et a1

'lnitiatives to Improve the Development of USAF CsI
Software,* MITRE, March 1984

I

50

%Appendix AS

-

Why Is Building Software Hard?

There are no radical breakthroughs now in view; moreover the very nature of software
makes it unlikely that there will be any - no inventions that will do for software
productivity, reliability, ald simplicity what electronics, traiistors, large-scale integration
did for computer hardware. We cannot expect ever to see two-fold gains every two years.
To see why this is so, and to determine what actions we must follow instead of hoping
tor breakthroughs, let us examine the difficulties of the software development process.
We divide them into essence, the difficulties inherent in the nature of the software, and
non-inherent probleima, those difficulties which today attend its production but which are
not inherent.
The non-inherent problems I discuss in the next section. First let us consider the
essence.
The esserce of a software entity is a construct of interlocking concepts: data sets,
relationships aunong data items, algorithms, and invocations of functions. This essence is
abstract, in that the conceptual construct is the same under many different representatiuus.
It is nonetheless highly precise and richly detailed.
I believe the hard part of building software to be the specification, design, a.d testing
of this conceptual construct itself, net the labor of representing it and testing the fidelity
of the representation. We still make syntax errors, to be s':r.,e; but they are fuzz compared
to the conceptual errors in most systems.
If this is true, building software will always be hard. There is inherently no magic.
Let us consider the inherent properties of this irreducible espence of modern software
systems: complexity, conformity, changeability, and invisibility.
Complexity
Software entities are mcre complex for their size than perhaps any other human
construct, because no two parts are alike (at least above the statement level). If they
are, we make the two similar parts into one, a subroutine, open or closed.

In this

respect software systems differ profoundly from computers, buildings, or automobiles,

where repeated elements abound.
Digital computers are themselves more complex than most things people build; they
have very large numbers of states. This makes conceiving, describing, and testing them
* This appendix is an extract from "No Silver Bullet", an invited paper presented in
Dublin by F. P. Brooks at the 1986 Congress of the International Federation of information
Processing. The full paper is in the Congress Proceedings.
51

%hard. Software systems have orders of magnitude more states than computers do.
Likewise, a scaling-up of a software entity is not merely a repetition of the same
elements in larger size, it is necessarily an increase in the number of different elements.
In most cases, the elements interact with each other in some non-linear fashion, and the
complexity of the whole increases much more than linearly.
Many of the classical problems of developing software products derive from thia
essential complexity and its non-linear increases with size. From the complexity comes
the difficulty of communication among team members, which leads to product flaws, cost
overruns, schedule delays. From the complexity comes the difficulty of enumerating much
less visualizing, all the possible states of the program, and from that comes the unreliability.
Computer programs do not break or wear out. The bugs one finds are either design flaws,
implementation errors, or are the consequences of changed environments and interface.
From the complexity of the functions comes the difficulty of invoking those functions,
which makes programs hard to use. From complexity of structure comes the difculty of
Extending programs to new functions without creating side effects. From complexity of
structure come the unvisualized states that constitute security trapdoors.
Not only technical problems, but management problems as well come from the
complexity. It makes overview hard, thus impeding conceptual integrity. It, makes it hard
to find and control all the loose ends. It creates the tremendous learning and ,inderstanding
burden that makes personnel turnover a disaster.
Conformity
Complexity alone is nothing unique to the software discipline. Phyaics deals with
terribly complex objects even at the "fundamental" particle level. The physicist labors
on, however, in a firm faith %hatthere are uuifying principles to be found, whether in
quarks or in unified field theorie.. Einstein repeatedly argued that there must eventually
be simplified explanations of nature, bet.,.-iae God is not capricious or arbitrary.
No such faith comforts the software engineer. Much of the complexity he must master
is arbitrary complexity, forced without rhyme or reason by the many human institutions
and systems to which his interfaces must conform. These differ from interface to interface,
and from time to time, not because of necessity but only because they were designed by
different people, rather than by God.
In many cases the software must conform because it has most recently come to the
scene. In others iv must conform because it is perceived as the most conformable. But
in all cases, much complexity cnmes from conformation to other interfaces; this cannot be
simplified out by any redesign of the software alone.

52

%Changeability
The software entity is constantly subject to pressures for change. Of course, so
are buildings, cars, computers. But manufactured things are infrequently changed after
manufacture; they are superseded by later models, or essential changes are incorporated
in later-aerial-number copies of the same basic design. Call-backs of automobiles are really
quite infrequent; field changes of computers somewhat leas so. Both are much less frequent
than modifications to fielded softwaze.
Partly this is because the software in a system embodies its function, and the function
is the part which most feels the pressures of change. Partly it is because software can be
changed more easily - it is pure thought-stuff, infinitely malleable. Buildings do in fact
get changed, but the high costs of change, understood by all, serve to dampen the whims
of the changers.
All successful software gets changed. Two processes are at work. As a software product
is found to be useful, people try it in new cases at the edge of, or beyond, the original
domain. The pressures for extended function come chiefly from users who like the basic
function amd invent new uses for it.
Succezsful software also survives beyond the normal life of the machine vehicle for
which it is first written. If not new computer3, then at least new disks, new displays, new
printers come along; and the software must be conformed to its new vehicles of opportunity.
In ehort, the software product is embedded in a cultural matrix of applications, users,
laws, and muchinc vehic!es. These all chonge continually, and their changes inexorably
force ch-.nge -upon the software product.
Invisibility
Software is invisible and unvisualizable. Geometric abstractions are powerful tools.
The floor plan of a building helps both architect and client evaluate spaces, traffic flows,
views. Contradirtcon* 'become obvious, omissions can be caught. Scale drawings of
mechanical paits and stick-5gure moaeis of molecules, although abstractions, serve the
same purpose. A geometric reality is captured in a geftmetric abstraction.
The reality of software is not inherently embedded in space. Hence it has no ready
geom'trkc reuresentation in tLe way that la.d has maps, silicon chips have diagrams, computers have cortriectivity schematics. As soon Ra we attempt to diagram software structure,
we find it to constitute not one, bat severan, general directed graphs, superimposed one
upon another The several grapi's may represent the flow of control, the flow of data,
patterna of dependency, time sequence, name-spaLe relatioiships. These are usually not
even planar, much lea hierarchical. Indend, one of the ways of establishing conceptual
co.,trol over such structure is to enforce link cutting until one or more of the graphs
becomes hierarchical [Parnas, 1979].
In spite of progress in restricting and simplifying the structures of software, they remain
inherently urvisualizable, thus depriving the mind of some of its moct powerful conceptual

%tools. This lack not only impedes the process of design within one mind, it severely hinders
communication among minds.
Past Breakthroughs Solved Accidental Difficulties
If we examine the three steps in software technology that have been most fruitful in
the past, we discover that each attacked a different major difficulty in building software,
but they have been the non-inherent, not the essential, difficulties. We can also see the
natural limits of each such approach.
Highi-Level Languages
Surely the most powerful stroke for software productivity, reliability, and simplicity has
been the progressive use of high-level languages for programming. Most observers credit
that development with at least a factor of five in productivity, and with concomitant gains
in reliability, simplicity, and comprehenfibility.
What does a high-level language accomplish? It frees a program from much of its
incidental complexity. An abstract program consists of conceptual constructs: operations,
data-types, sequences, and cornmmunication. The concrete machine program is concerned
with bits, registers, conditions, branches, channels, disks, and such. To the extent that
the high-level language embodies the constructs one wa~nts in the abstract program and
avoids all lower ones, it eliminates a whole level of complexity that was never inherent in
the program at all.
The most a high-level language can do is to furnish all the constructs the programmer imagines in the ab'stract program. To be sure, the level of our sophistication in
thinking about data structures, data types, and operations is steadily rising, but at an
ever-decreasing rate. And language development approaches closer and closer to the
sophistication of users.
Moreover, at some point the elaboration of a high-level language becomes a burden
that increases, not reduces, the intellectual task of the user who rarely uses thc esoteric
constructs.

I

Time-Sharing

Most observers credit time-sharing with a major improvement in. the. productivity of

programmers and in the quality of their product, although not so large as that brought by
high-level languages.

Time-sharing attacks a quite different difficulty. Time-sharing preserves immediacy,
and hence enables one to maintain an overview of complexity. The slow turnaround of
batch programming means that one inevitably forgets the details, if not the very thrust,
of what he was thinking when he stopped programming and called for compilation and

54

%execution. This interruption of consciousness is costly in time, for one must refresh. The
most serious effect may well be the decay of grasp of all that is going on in a complex
system.
Slow turn-around, like machine-language complexities, is an unnecessary rather than an
essential difficulty of the software process. The limits of the contribution of time-sharing
derive directly. The principal effect is to shorten system response time. As it goes to
zero, at some point it passes the human threshold of noticeability, about 100 milliseconds.
Beyond that no benefits are to be expected.
Unified Programming Environments
Unix and Interlisp, the first integrated programming environments to come into
widespread use, are perceived to have improved productivity by integral factors. Why?
They attack the incidental difficulties of using programs together, by providing integrated libraries, unified file formats, and pipes and filters. As a result, conceptual
structures that in principle could always call, feed, and use one another can indeed easily
do so in practice.

WM
3W

This breakthrough in turn stimulated the development of whole toolbenches, since each
new tool could be applied to any programs using the standard formats. How much more
gain can be expected from the exploding researches into better programming environments?
One's inst.inctive reaction is that the big-payoff problems were the first attacked, and have
been solved: hierarchical file systems, uniform file formats so as to have uniform program
interfaces, and generalized tools. Language-specific smart editors are developments not
yet widely used in practice, but the most they promise is freedom from syntactic errors
and simple semantic errors.
Perhaps the biggest gain yet to be realized in the programming environment is the use
of integrated database systems to keep track of the myriads of details that must be recalled
accurately by the individual programmer and kept current in a group of collaborators on
a single system.
"M
N2

~Surely
this work is worthwhile, and surely it will bear some fruit in both productivity
and reliability. But by its very nature, the return from now on must be marginal.
Conclusion
All of the technological attacks of the software process are fundame ntally limited by
the productivity equation:

time of task

=

3(frequency)j x (time)i

If, as I believe, the conceptual components of the task are now taking most of the time,
*

-- - - - - - -- - - - - - - -- - - - - --

55

%then no amount of activity on the task components that are merely the expression of the
concepts can give large productivity gains.
We are left with the inherent task - getting the complex concept. right, and changing
them correctly as the world keeps changing about them. This is a human activity, and a
ltabor-intensive one.

56

%Technical Report
8E1-6-TR-2
September 1986

Proposal for a New "Rights In Software" Clause
for Software Acquisitions by the Department of
Defense

Pamela Samuelson

Appt wed for public release.
Distrbtlon unlimited.

Software Engineering Institute
Carnegie Mellon University
Pittsburgh, Pennsylvania 15213

%Table of Contents

1I Background1
2. Issues3

2.1 Should DoD Adopt a "Standard Rights In Software Clause"
and Take Software Out of the Technical Data Rights Clause?
2.1.1 Resomns that Support a Separate "Rights InSoftwre"Polkcy
2.1.2 Reasons not to Adopt a Separate Software Acquisition Clause
2.1.3 Conclusion
2.2 What might a Standard Rights In Software Clause Look Uke?
2.2.1 The M,4del Standard Rights In Software Clause
2.2.2 Commeltry to the Model Standard Rights In Software Clause
2.3 It DoD Does Not Adopt a Separate Rights In Software Clause,
How Should It Revise the Standard Data Rights Clause to Improve
its Software Acqi~sltion Practices?
2.3.1 Definitions
2.3.2 Policy as to Publicly Funded Software
2.3.3 Policy as to Privately Funded Software
3. Conclusion

58

I V

Ir-

rlu

i'

~

.

~.~~

k "C~~
VIV. V WN LU~\.N1LV

bU%1 V

.I

LP%<

X.X

3
3
5
7
7
7
10
13
13
15
17
21

%Proposal for a New "Rights In Software"
Clause for Software Acquisitions by the
Department of Defense
Abstract
This report recommends three distinct regulatory strategies for addressing difficulties the
Department of Defense (DoD) has been experiencing with respect to legal Issues related
to software acquisitions. First, the report reiterates the Software Licansing Project's earlier recommendation that the DoD adopt tne proposed Federal Acquisition Regulation
(FAR) data rights provisions instead of the proposed revisions to the DOD supplement to
the FAR (DOD FAR SUPP).
Secondly, in the event that the Defense Department chooses to adopt a data rights
procurement policy different from that found in the data rights provisions of the proposed
FAR, this report recommends that the DOD adopt a separate "Rights in Software' clause
for software acquisitions, rather than continuing the present practice of handling software
procurements under the "Rights in Technical Data" clause. Reasons in support of a
separate software acquisition policy, as well as a beginning model "Rights in Software"
clause are offered.
Finally, in the event that the DoD elects to retsin the procurement format presently found
in the DOD FAR SUPP provisions governing software and technical data acquisitions,
this report offers several concrete recommendations for changes to those regulations
*which
should result in a procurement policy which more effectively meets the mission
needs of the Defense Department.

1. Background
The Softwaro Licensing Project (SLP) of the Software Engineering Institute (SEI) has written two
previous reports on the Department of Defense's (DOD) software acquisition policy. The first of these
reports was 'Toward a Reform of the Defense Department Software Acquisition Policy,, CMU/SEI-86TRI [2] (hereinafter referred to the "First Report'). It surveyed a range of problems that DOD personnel had identified as software licensing problems currently being experienced by DoD. One chapter
of the First Report was devoted to an analysis of the data rights regulations that govern acquisitions
of software by DoD. The First Report concluded that a substantia: revision of DOD's standard data
rights clause would be desirable.
The second SLP report was "Comments on the Prcpused Federal and Defense Acquisition
Regulations," SEI-86-*,TM2 [1] (hereinafter referred to as the 'Second Report'). It recommended that
the Department of Defense adopt the proposed Federal Acquisition Regulation (FAR) data rights
provisions instead of its proposed revisions to its supplement to the FAR data rights regulations. The
Second Report made this recommendation for four reasons: (1) The proposed FAR data rights
regulations present a more concise and comprehensible regulatory scheme than either the current or

59

%proposed DoD regulations. (2) The proposed FAR data rights policy is also more compatible with
standard software commercial practices and provdes more incentives for industry to make its best
technology available to the government than does the DOD policy. (3) At the same time, the proposed FAR data rights pol;cy would give to the government a number of rights that DoD would seem
to need to fulfill its mission (including rights which the current and proposed DoD regulations fail to
claim for DoD). (4) Both statutory and policy reasons support having a uniform set of federal data
rights regulations rather than having two policies, one for DoD and one for all other federal agencies.
This report is the third SLP Report to concern itself with the DoD procurement regulations affecting
software. While we continue to stand on our recommendation that DoD adopt the FAR data rights
provisions, we understand that for various reasons, the Department of Defense may find it undesirat,'e to adopt the proposed FAR data rights policy and may decide to continue with its separate data
rights policy.
In the event that DoD chooses to continue its separate approach to software acquisitions, we would
have the Department of Defense consider three further recommendations which are set forth in this
report. First, we recommend that the DoD create a separate "standard rights in software clause', that
is, to break software out of the standard technical data rights clause. Some part of the reason why
DoD has experienced so much difficulty in its software acquisition policy is, we believe, due to the
quasi-technical-data-rights orientation o! its present policy, an orientation which is inappropriate for
software acquisitions.
Second, we offer a draft standard "rights in software" clause fo., DoD's consideration. This clause
provides for separate treatment of software acquisitions, d-stinct from that accorded technical data
under the standard data rights clause. This "rights in software" clause presents several unique features which distinguish it from the standard data rights clause. These include: the inclusion of
software documentation wifhin the definition of the term "software," the establishment of government
purpose rights as the standard "ceiling" of rights that the government obtains in publicly funded software, and the provisicon that software will retain its restricted rights status even when slight modifications are made at the request of the government.
Third, in tVa event that DoD chooses not to adopt our first two recommendations, and decides to
retain the basic structure and content of the existing standard data rights clause, there are still a
number of specific changes to that clause, as it affects software, that we believe would be in the
government's best interest to adopt. There are 22 specific recommendations for changes to the text
of the DoD standard data rights clause discussed within, all of which would, in our view, improve
DoD's software acquisition process.

2

60

%2. Issues
2.1. Should DoD Adopt a "Standard Rights In Software Clause" and
Take Software Out of the Technical Data Rights Clause?
For well over a decade, DoD has acquired rights In software by means of the same standard clause
as that used to acquire rights in technical data (DoD FAR SUPP sec. 52.227-7013, also known as the
standard data rights clause, referred to hereinafter as "SDRCO). We understand that the Department
Is currently considering adopting a separate clause for its acquisitions of rights in software, that Is,
breaking software out of the technical data rights provisions of the SDRC. Although we believe that
the Department can have a substantially improved software acquisition policy without such radical
surgery to the SDRC (after all, we have recommended adoption of the FAR data rights policy which
retains a unified technical data and software policy), we believe that, on the whole, tie Department
would be well served by making the change to a separate rights in software policy for the reasons
discussed below.

2.1.1. Reasons that Support a Separate "Rights In Software" Policy
2.1.1.1 The current DoD policy already partially dlfferenthites software from technical data.
Although DoD has long had a policy of acquiring rights in software under the same SDRC that is used
in acquisitions of rights in technical data, software has for some time been partially differentiated from
technical data within the body of the SDRC. The most obvious difference is in the rights the government takes as a matter of course in privately developed software, as compared with privately developed technical data. Software's "restricted rights" are very restrictive (e.g., to particular computers)
as compared with technical data's "limited rightsV which permits use or copying throughout the govemnment. This reflects that the Department has already recognized that software and technical data
a\&I different. The SDRC also recogrizes that the rights that the government needs in software, and
the limitations that are reasonable for industry to impose on the government's rights in software are
different from those that pertain to technical data.
The question we have beer, raising is whether software is differentiated enough in the SDRC and
differentfated in the right ways. For various reasons discussed in our First Report, we believe that
DoD has not yet adequately differentialted between tecnnical data and software. This is why, we
beliove, derivative works rights which are critically important as to software, have been omitted from
the technical data oriented SDRC, which defines unlimited rights without reference to a right to make
derivative works. A separate software clause would facilitate appropriate differentiation between
software and technical data.
2.1.1.2 Economic reasons why software documentation should be treated differently from
technical data.
The function and purpose of software Is different from that of technical data. Software performs
tasks; technical data merely conveys information. Because of this, the economics underlying the
development and marketing of software and technical data are significantly different. Software gener

3

61

...

%ally involves significant research and development costs which can only be reouped through the
marketing of the product, software Itself, whereas tevhnical data Isgenerally produced as an ancillary
step Inthe process leading to production of the actual Item to be marketed.
The critical point here Isthat the capital cost of design and development (including the cost of software tools andlor CAD/CAM programs which aided Inthe development effort) are recouped as part of
the sale of the system, niot through sales of technical data that might have been generated Indeveloping the system. DoD's policy with respect to hardware systems takes this into account by treating
hardware systems in a manner different than It treats technical documentation. DoD's present policy
with respect to software, however, Is heavily technical data oriented, and does not allow software
design costs to be recovered Inthe same manner.
Thus, the economics of software development Indicate a need for breaking software (anid the documentation which is an Integral part of Its development and evolution) out from the quasi-technical data
treatment it has thus far received. With regard to development costs and capitalization, software is in
many ways more like a hardware component than It is like the technical documentation which supports the hardware. The DoD procurement policy needs to be structured so as to take account of
these technical and economic similarities between software and hardware, as well as the dissimilarities between software and technical data.
This policy should also recognize that unlike hardware, software is an evolutionary product - that Is,it
is in a state of constant development as maintenance and enhancement work is continually done to
improve upon and/or alter the functioning of the software. As an evolutionary product, the documentation supporting the software is in fact a critical part of the software product Itself. For this reason,
the software documentation should be treated in the same manner as the executable version of the
program. A property structured software acquisition clause can accomplish this.
2.1 .1.3 Outside of the DoD regulations, different Intellectual property rights may attach to
software than to technical data.

Pei

Software is a unique intellectual property in that it can be protected under the copyright law, trade
secret law, and patent law. The unique nature of software allows itto be copyrighted without revealing all of its "secrets" which means that trade secret and copyright protection can coexist inthe same
subject matter. It is rare for a firm to copyright technical data that the firm wanted to claim as a trade
secret, because the Copyright Office generally makes any deposited work available for public Inspec.tion and copyright law treats such things as manufacturling instructions or engineering designs as
"Ideas" which are in the public domain. Firms tend to keep manufacturing Instructions and other
technical data solely as trade secrets. A separate clause to govern software acquisitions could take
into account differences inintellectual property protection affecting software and technical data.
2.1.1.4 The educational value of a separate software clause.
A new clause to govern software acquisitions could accomplish a break with the past, and engender a
move away from the quasi-data rights orientation which has pervaded software acquisitions. A new
clause could pave the way to a now *mind set" for those who work in the area of software and data
rights acquisitions. Such a clause would provide a point of departure for Te-educating procurement
4

62

%personnel regarding the nature of software. In this way, It could create a fresh way of viewing
software acquisitions, one more in line with the economic and technological realities of the software
industry.
2.1.1.5 Improving relations with Industry.
It is unfortunate that relations between the software industry and the Department of Defense are at
present somewhat strained over software data rights issues. Many industry representatives seem to
feel that DoD software procurement policy is confiscatory. The adoption of a separate clause to
govern software acquisitions, which would break such acquisitions out from the policies with which
industry has been unhappy, could go far to improve government-industry relations. At the very least,
the perception that DoD is making some effort to alleviate the areas of conflict with industry could be
valuable in this regard.

2.1.2. Reasons not to Adopt a Separate Software Acquisition Clause
2.1.2.1 The overlap between software and technical data.
A separate software clause is not necessary to significantly improve the DoD's software acquisition
policy. Even we conclude that the FAR data rights policy, which retains a unified approach, would be
an excellent policy for DoD. This is one reason not to break software out of the technical data clause.
There are others as well.
There is, for instance, some artifice in the distinction between software and technical data. Technical
data car be incorporated into a computer data base, for example, which would seem to transform it
"into software. in fact, virtually anything that can be written on paper can be transformed Into a
machine readable form. The DoD would need to sort out the computerized technical data problem
which its present regulat*.ons also fail to do but apart from this, software and technical data are
sufficiently distinct that a separate policy is appropriate, as DoD's present SDRC a:ready demonstrates.
2.1.2.2 Would DOD seem to be "caving In" to Industry if It adopted a separate software
clause?
Since software resembles technical data and hin, long been treated within the technical data policy,
and since the software industry has been lobbying for a special software policy, one problem that
DoD may see with a separ- a software clause is that it may appear to some that the DoD would be
too generous to industry, especially if the Deportment allows industry to retain greater rights in software than in technical data. DoD's response to such charges should, however, be that the d;ffererntial
treatment of software would actually save the government money in that the government would not
be forced by the regulations into purchasing the more expensive "govemment-wide rights" to software
documentation in those instances where a site license is adequate to the needs of the government
and that better software at lower development costs will be made available to the government if it
provides better incentives to the software industry. Such responses should serve to silence the
critics.
2.1.2.3 The iseed to retrain DoD's contracting personnel as to any new software clause.

oil"

63

%A separate rights clause to govern software acquisitions has the potential to further complicate the
DoD acquisition process. Those who have long experience with the SDRC have become used to
muddling through the present system. They would have to be retrained about rights in software, and
this is no small job.
The DoD needs, like private Industry, to be Involved In the evolution of a conceptualization of software
and software acquisition which Is consistent with the technological, economic and legal realities of
software development. A separate treatment for software, along with the retraining which would need
to be undertaken in conjunction with such a change, could go a long way toward developing a new
and more dynamic conceptual framework for dealing with software.
2.1.2.4 The desirability of an overhaul of the DoD procurement policy as to Intellectual
property.
The DoD would benefit greatly from a more substantial overhaul of the procurement regulations to
make them more compatible with traditional and newly developing intellectual property law. A more
integrated, more unified intellectual property policy could bring together DoD's policies as to
copyright, patent, semi-conductor chip design, trade secret and trademark law. Advances in new
technologies are bringing together and blurring the the lines between these traditional forms of intellectual property protection. As the new technologies continue to advance, the need to integrate
policies in these areas will become more acute. Additionally, government attorneys working In the
software/data rights area must of necessity have some grounding in the traditional forms of intellectual property law. Given this, it seems wise for DoD to draw upon the knowledge and expertise
already possessed by its lawyers involved in this area by making its policies consistent with the
already existing body of intellectual property law.
A separate clause for software acquisitions will contribute to a fractionated rather than a unified
system of intellectual property regulations. The time and energy expended in adopting a separate
software acquisition clause would probably be at the expense of efforts which might othewise have
been invested in developing a broader. more integrated Intellectual property policy for the department, a policy which needs generally to be more integrated with copyright and trade secret law.

2.1.3. Conclusion
On the balance, we believe that the advantages presented by a separate software acquisition clause
outweigh the potential disadvantages. We would recommend, therefore, that the DoD adopt a software acquisition clause as part of its procurement regulations. A suggested model clause Is included
in this report. It should be noted that the clause, while offering a fresh approach to software acquisition, only touches briefly on software maintenance and enhancement. In recognition of the critical
importance of these issues, the next phase of this project's research will focus specifically on these
issues. A more in-depth treatment of maintenance and enhancement will be forthcoming with the
project's next report.

6

64

%2.2. What Might a Standard Rights In Software Clause Look Like?
2.2.1. The Model Standard Rights in Software Clause
(a) Definitions
As used in ,his clause, the following terms have the following meanings:

government purpose
the fulfillment of a legitimate federal government function, Including uses or disclosures for competitive reprocurements and maintenance and enhancement purposes; the term includes disclosure to and use by other contractors and any
state, local or foreign government where such disclosure or use will fulfill a
legitimate federal government purpose; the term does not Include a general distribution of the software to defense contractors or other more limited distributions of
the software that may have a significant negative etiect on the commercial market for such software. Nor does it include a disclosure that permits the recipient
to disseminate the software without restriction or to develop software for nongovernmental
sales in competition with the owner of intellectual property rights in
it.
government purpose license
a license to the federal government that grants the government rights to use,
duplicate, disclose, distribute, prepare derivative works, and publicly display software for government purposes, and to authorize others to exercice such rights
when doing so will fulfill a legitimate federal governmental function. When software provided to the government by one contractor is distributed or disclosed by
the government to a subsequent contractor for a government purpose, the subsequent contractor shall be bound by the terms of the govemment purpose license.
restricted rights license
a license to the federal government that at a minimum grants the government
rights
(1) to use software in the computer for which the software was acquired;
(2) to use software in a backup computer ifthe computer for which it was
acquired becomes inoperable;
(3) to make copies of the software necessary for backup and reverse
engineering purposes;
(4) to adapt and modify the software; and
(5) to authorize support contractors to exercise the rights described In (1)
through (4), subject to the same restrictions as bind the government.
restricted rights software
software that has been developed at private expense, including software as to
which only slight modifications are made to adapt it for the government needs
with public funds. The term "developed" means fixed In a tangible medium of
expression. The term "at private expense" means entirely funded by the contractor and without any government reimbursement, direct or undirect other than
through IR\&D co=t allocations.
software
computer programs, computer data bases, and documentation pertaining thereto
Including but not limited to such programs in any machine readable printed or
interpreted form, system reference manuals and user manuals.

7
65
CM

wu~

n

h

4

~~

A

~J

1LN

A

r

~iU

~

*

~

J

M~N

.

%(b) Rights of the Govemment
(1) Public Domain Software: Thero shall be no restrictions
on the government's right to use, duplicate, disclose, distribute, display or make derivatives of software that is in the public domain.
(2) Government Purpose Liconses: The government shall hav4e a government purpose license in all software deliverable under this contract that was developed at public expense. The
government may also negotiate to obtain a government purpose license in software that was developed at private expense.
(3) Restdicted Rights License: The government shall have a resticted rights license in all
restricted rights software deliverable under this rontract. Written permission of the owner of such
software will be required before the government may make or authorize other uses or disclosureas of
this software.
(4) Neautiating for Additiona' RiarL: The gova.mment may negotio-te to obtain more rights
in restricted tights sotiware than the five stanidard rights that are narried in the definit;on of the
restricted rights license. Additiona;;y, the government and contractor may nnfliate to define the
uses the government may make of software within the scope of Wie government p, (pose license.
(5) inco.]oration of Other Sofiware: When a contractor incoiporates into software to be
delivered to the government modules or subroutines in which the contractor does not own all intllectua! property rights, th'j contractor shall obtain for Ihe government at least a restricted rights license in such incorporated modules or subroutines.
(6) iohts from Sut-contractors: The govemment shall have the same minim•in rights in
software developed by subcontractors as in software deve!oped by primo contractors.
(7) Cha.enaing Restrictive Leaends
legends.
tive

The government may challenge inappropriate restric-

(c) Rights of ContVantors and Subcontactore
(1) Ownership: Unless the special works clause has i)een invoked, whoever develops software deliverable under this contract shal! be considered the owner of all intellectual property rights in
it, subject to a restricted rights or government purpose 1,cen3e to the govemment as provided in
Section (b).
(2) Restrictive Markings: The contractor or subcontractor who owns intellectual property
rights in software may attach appropriate restrictive markings to the software in accordance with this
clause.
(3) Direct Delivery to the Government: Subcontractors under this contract may deliver
restricted rights software directly to the government rather than to the prime cotrActor unless the
software is needed by the prime contractor for installation in the system that the contractor is required
to deliver to the govemrme.t.

S8
66

%(4) _No Leverage: Neither the prime contractor nor any Intermediate subcontractor shall use
its power to award subcontracts as a means of acquiring greater rights In software from its subcontractors than is needed to perform 1he government contract.
(5) Flowdown to Subcontractor Whenever any software is to be obtained from a subcontractor under this contract, the contractor shall use this same clause in the subcontract, without alteration.
No other clause shall be used that will enlarge or diminish either the government's or the contractor's
rights in the subcontractor's software which is to be delivered to the government.

(d) Restrictive Legends
(1) No Marking If In Public Domain: Software that is in the public domain shall be delivered
with no restrictive markings.
(2) Government Purpose Rights Legend: Software in which the government has government purpose rights is to be delivered to the government with the following restrictive legend:
Government Purpose Rights
Property of: (contractor or subcontractor's name)
Standard Restricted Rights Legend: Restricted rights software in which the government has only the
standard five minimum rights are to be delivered to the government with the following restrictive
legend:
Restricted Rights
Property of: (contractor or subcontractor's name)
(4) Other Restricted Rights Legend: When the government and the contractor (or
subcontractor) have negotia!ed an arrangement whereby the government will get more than the standard five inir.murn rights in restricted rights software, the software shall be delivered with the following re;trictive !egend:
Expanded Restricted Rights
Property of: (contractor or subcontractor's Name)
Contract No:
(5) Coovriaht Notices: Unless the special works clause has been Invoked, the owner of
intellectual property rights in software may attach appropriate copyright notices to software delivered
under this contract.

2.2.2. Commentary to the Model Standard Rights In Software Clause
There are a number of respects in which this standard rights in software clause differs from the
"- C, among them:
* that software is defined to Include documentation;
e that governmental purpose rights are the standard "ceiling" of rights that the government
has in publicly funded software;

I
67

9

%"*that there is no differentiation in the level of the govemnment's rights dependent on
whether or not the contractor copyrights the software;
"*that the government will have a right to prepare, or authorize prepar~thon of, derivative
software from software developed at public expense;
"*that software will niot lose its restricted rights status If only slight modifications are made
to it at the request of the government;
"*that use by support contractors (subject to restrictions binding the government) Is Included Inthe set of restricted rights;
"*that *developed* is defined in a manner more consistent with copyright than patent standards;
"*that no explicit reference Is made as to the contractor's right to claim a copyright because
we regard this as implicit Inthe clause's recognition of the developer's night to Intellectual
property rights in the software.
Before discussing some of these features, it may be helpful to describe the circumstances In which
we would envision this clause being used.
2.2.2.1 The quasi-mandatory nature of the standard clause.
The SDRC is required to be inserted In all Defense Department software acquisition contracts. The
present SDRC contemplates two situations in which the government's rights in the software may be
different than those that the SDRC itself prescribes:
1. When the government uses the special works clause in a software development contract, and
2. When the contractor and the government negotiate an agreement giving the government more than the four standard minimum rights in privately developed software.

I
Ar

The SDRC will govern all rights In software matters unless one of these circumstances is present.
Our proposed standard software clause would operate in much the same fashion. That Is,itwould be
a mandatory clause for insertion into all DoD software acquisition contracts unless one of a set of
authorized alternate rights acquisition clauses was used in the contract. We would recommend retention of the two already authorized alternatives, and would recommend serious consideration of two
other authorized alternatives, one permitting the government to negotiate for less than government
purpose rights when there is substantial private funding of the software's development in addition to
some public funding, and another for acquiring less than the standard set of minimum rights In software tools and CAD/CAM programs.

2.2.2.2 A
"mixed funding" alternative to equitably distribute rights based on public and

private funding.
As one alternative to the standards "rights In software" clause, the DoD should consider adopting a
clause which would equitably allocate rights in software In mixed funding situations. The DoD Authorization Act of 1985 seems to contemplate adoption of a data rights policy that differentiates be*tween wholly government funded and partly government funded projects. DoD's present regulations
*have not responded to this Congressional directive. The DoD would, of course, need to address
issues regarding what forms of contribution to a project constitute private funding (resources or cash).
what degree of private funding would be necessary to trigger the mixed funding alternative, how much
flexibility to allow contracting personnel in structuring mixed funding arrangements, and the like.

%2.2.2.3 An alternative clause to obtain less than the standard minimum rights In software
tools and CAD/CAM programs.
Additionally, the DoD might consider adopting another alternative allocation of rights clause, one
which would allow the DoD to obtain less than minimum rights in certain items such as software tools
and computer aided design/computer aided manufacturing (CAD/CAM programs). Since software
tools and CAD/CAM programs are such valuable resources of private firms, contractors are loath to
provide these tools to the government under the standard rights arrangements. It would seem that
DoD would be wise to provide in its regulations the flexibility to negotiate for some access to these
items, on the theory that partial access will in some instances be better than none at all. It is in DoD's
interest to assure contractors that they can provide their best technology to the DoD without tear of
loss of these rights in their software.
2.2.2.4 Why government purpose rights Is the standard celling of rights under the clause
Instead of unlimited rights.
As our First Report has indicated, it seems that under the standard data rights clause the government
now obtains government purpose rights rather than unlimited rights in publicly funded software in
which the contractor claims a copyright. It is not clear why the government has chosen to provide this
incentive to contractors to copyright software. After studying this matter, we have concluded that
there should not be a difference in the extent of the government's rights depending on whether the
software is copyrighted by the contractor. Because it appears that the government is already willing
to accept government purpose rights for copyrighted software developed at public expense, we believe it is reasonable for the government to use the same policy as to all publicly funded software.
Indeed, we fail to see why the government would ever need more than government purpose rights in
publicly funded software.
2.2.2.5 The definition of the term "developed" should be grounded In principles of copyright
law.
The approach DoD has taken toward defining "developed* within the meaning of "developed at
private expense" has been a patent-oriented definition of the term. Indeed, the government's patent
lawyers seem to have diligently and aggressively attempted to use a patent standard toward software
development so as to establish for the government as broad a set of rights as possible in software.
As discussed in the First Report, one result of claiming this broad set of rights for the government has
been to create significant disincentives for contractors to deliver their best technology to the government.
The model clause takes a more copyright-like approach to defining "developed." Because software Is
copyrightable, and copyright law allows intellectual property rights to attach whenever a work is fixed
in a tangible medium of expression, it seems appropriate for the government regulations applicable to
software to be more consistent with this body of intellectual property law (which Is,after all, the most
Important body of federal intellectual law affecting software). (Although software may sometimes be
patentable, software patents are much rarer than software copyrights.) A copyright approach to a
definition of "developed" would also be more consistent with the nature of the software development

69

04%

1

%process. Unlike hardware, software Is almost continually In the process of development. Copyright
law which Is attentive to this evolutionary nature of software, Is more appropriate than a patentoriented standard.
We recognize that because software Is a hybrid, lying somewhere between traditional copyright and
patent subjeel matters, it Is difficult to find the appropriate location on the continuum as to when
software is "developed" or not developed. The proposed DoD regulatory standard would seem to call
for software to have gone through extensive testing before it can be deemed developed. We consider this to be one extreme of the continuum. The "fixed In a tangible medium" standard which we
have chosen to include in the model clause may represent the other extreme.
In shoosing this standard, we were deferring to the copyright law since that is the nearest bf., of
intellectual property law applicable to software. We offer this definition as a point of discussion, and
understand that DoD may prefer a mo:e operational definition. As a viable alternative to the definition
we have presented, the DoD might consider a compromise between the copyright approach to the
definition of *developed" and an operational definition which does not require the developer to go to
an extensive degree of testing before software can be deemed developed. It is important that such a
definition recognize that software Is in a state of continual development and Improvement which
makes impractical any definition which focuses on finished products. This conflict points out the
predicament encountered by government and industry alike In dealing with this strange hybrid subject
matter. To the extent software is like hardware, it would seem an appropriate subject matter to hold
to the higher, more operationally oriented standard of development under the patent law, and to the
extent it is like technical data and is subject to continual modification, it seems more appropriate to
the more flexible standard for development found in the copyright law. This is a dilemma, but DoD
has already tried unsuccessfully to adopt a patent standard for defining "developed" and found the
software industry to be so hostile to it that another approach must be found.
..2.2.6 Respects In which the model standard rights In software clause Is more advantageous
to the DoD than the SDRC.
In addition to the benefits the DoD would realize as a result of eliminating disincentives which cause
some developers to withhold their best technology from the DoD, there are several respects in which
the model standard rights in software clause gives to the DoD broader rights than those which it
would acquire under the present treatment of software acquisitions under the SDRC. These Include:
* the right to reverse engineer as a minimum right in software acquisitions;
* the right to license support contractors as a minimum right in software acquisitions;
° the right to make derivative works as an explicit part of the government purpose rights
package;
e a very broad definition of government purpose rights which includes such rights as use or
disclosure for competitive reprocurements, as well as disclosure to and use by state,
local and foreign governments.

12
7

%2.3. If DoD Does Not Adopt a Separate Rights In Software Clause,
how Should it Revise the Standard Data Rights Clause to
Improve its Software Acquisition Practices?
Sections 1 and 2 ol this report detail the reasons why a separate software clause may be in the
DoD's best interests and then sets forth a model software rights clause for the Department's consideration. In the event the Department of Defense has not been convinced of the desirability of
taking this approach, there is still much that can be done to improve the existing SDRC as it affects
software. The following 22 recommendations are distillations of many of the points made in the First
Report of the SLP. (Page and chapter numbers in parentheses below refer to the First Report.)

2.3.1. Definitions
2.3.1.1 Don't overdefine software terms.
Six software-related definitions are included in the SDRC. Only three seem to be significant in the
body of the standard data rights cdause -- software, software documentation, and commercial software. Only these three need to be defined. Also, the SDRC speaks constantly of "computer
software" when it is only necessary to say "software", because 'computer" is already included In the
software definition.

4

2.3.1.2 If thp distinction between commercial and other-than-commercial software Is to be
retained, provide a more precise definition of what Is meant by commercial computer
software.
The SDRC provides for two different sets of restricted rights applicable to privately developed software, one for "commercial" software and one for other software (or commercial software whose
owner opts to have it treated as other-than-commercial software). (Different restrictive legends are
supposed to be attached to software, based on what kind of software is to be delivered.) Unfortunately, the existing definition of "commercial computer software" is so vague as to be a poor guide
as to what software will qualify for commercial restricted rights treatment (see pp. 23-4).
2.3.1.3 If two sets of restricted rights for privately developed software are retained, the definitional section of the clause should Include and define both sets of restricted rights.
As noted above, there are two categories of privately developed software which are presently subject
to different sets of restricted rights. The definitional section of the SDRC sets forth only one definition
of restricted rights, which a later section of the SDRC seems to make applicable only to other-thancommercial software. The other set of restricted rights, those applicable to commercial software (and
its documentation), are not set forth until subsection (b)(3)(ii). In order to achieve consistency, these
"commercial restricted rights" should also be set forth in the definitional section of the clause. (p. 26.)
2.3.1.4 Define what Is meant by "government purpose," perhaps clarifying Its meaning by
providing some examples.


%DoD policy allows a contractor to copyright any software developed under a government contract
(unless it is a "special work'). Subsection (c) of the SDRC provides that the contractor must grant to
the government a copyright license "for government purposes" as to any work in which he has taken
a copyright. However, there is no definition of *government purpose,' either in that subsection or In
the definitional section. This omission creates uncertainty as to the extent of the government's rights
in publicly furnded copyrighted software (see pp. 6, 24-5, and Chapter 7).
2.3.1.5 Expand the definition of unlimited rights to Include the right to prepare derivative
works.
The present SDRC definition of unlimited rights fails to make explicit whether the government will
have the right to prepare derivative works when it has unlimited rights In software. Such a right is
particularly important as to software because maintenance, enhancement, reuse, translation, rehostIng and retargeting are all dependent on having such a right (see pp. 19, 54, 72). The fact that that
the proposed Fedetal Acquisition Regulations (FAR 52.227-14(a)) would give other governmental
agencies a derivative works right in unlimited rights software •;,.uid weaken DoD's argument that the
derivative works right is Implicitly Included in its unlimited rk[hts policy. In light of the Importance of
this right to DoD, it would seem prudent for DoD to take the precaution of including the derivative
works right within its unlimited rights.

2.3.2. Policy as to Publicly Funded Software
2.3.2.1 Clarify that unlImited rights is a kind of license, not an ownership right.
The project's research revealed that DOD personnel had at least four different interpretations of the
meaning of unlimited rights vis a vis ownership rights.. Intellectual property law would likely treat
"•unlimited rights" as a broad license, not as an ownership interest. In order to avoid future misunderstandings and possible litigation, this concept needs to be clarified (see pp. 24-25, Chapter 7).
2.3.2.2 Clarify DoD's intent as to the effect a contractor's claim of copyright In publicly
funded software will have on the government's rights In publicly funded software.
There is an ambiguity in the present SDRC concerning the extent of the government's rights in
copyrighted software d3weloped at public expense. One part of the SDRC seems to give DoD unlimited rights in it because it was developed at public expense and another part gives the government
only govemmert purpose rights if the contractor decides to retain a copyright in the software. DoD
should clarify its Intent on this matter.
2.3.2.3 If DoD decides to retain the apparent policy of allowing a contractor's copyright to cut
back the government's unlimited rights license to a government purpose license, It should
require the contractor to give DoD early notice of his Intent to claim copyright.
A further disadvantage of the present SDRC as regards contractor copyrights in publicly funded
software is that it appears that the government will typically not know the extent of its rights - whether
unlimited rights or government purpose rights - until the software is delivered to the government, that
is, until It sees whether the software was delivered with or without a copyright notice attached. The
government may want to require notice of an Intent to claim copyright at the time the contract is
entered Into so that it can plan accordingly.
14

7'r.

%2.3.2.4 Revise the special works clause so that DoD will be able to take broader rights in
software when It needs them.
The DoD's special works clause (DFARS 52.227-7020) purports to claim a direct copyright for the
government under the "work for hire" doctrine. This clashes with Section 105 of the Copyright Act (17
U.S.C. Sec. 105) which prohibits the government from taking direct ownership rights in copyrighted
works. Use of the current special works clause would seem to have two effects: (1) to preclude the
contractor from claiming a cop))TIht in the software and (2) to put the software into the public domain,
since neither the government nor the contractor can own it.
Since copyright law does permit the government to own copyrights by assignment, a copyright strategy similar to that adopted by NASA and proposed for the FAR should be considered by DoD. (p. 21,
Chapter 5.)
2.3..5 DoD should either give up Its claim of unlimited rights In non-deliverable software or
make a deferred ordering clause standard.
The SDRC seems to give the government unlimited rights in several categories of software, although
their delivery may not be required by the contract (SDRC (b)(i).) Without the inclusion of a deferred
ordering clause, it appears that the government would not have the right to require delivery of any of
this non-deliverable software. The existence of this unenforceable inchoate right only serves to
frustrate both the government and industry.
We recommend that DoD examine whether it needs to claim unlimited rights in these nondeliverables. If it is decided that such a right is needed, a deferred ordering clause should be made a
standard part of the contract (see pp. 19-20).
2.3.2.6 In "mixed funding" situations, (I.e., where both public and privaie funds are used to
develop the software DoD should provide an option for the government tc take less than unlimited
rights.)
This would provide needed incentives to software firms to invest some of their own capital in software
development which could result in a higher quality product and in lower initial acquisition costs. it
would also conform with the apparent congressional intent reflected in Section 2320 of the Department of Defense Authorization Act of 4985, (Public Law 98-525, 10 U.S.C. Sec. 2301, 2320.)
One possibility would be to give the government unlimited rights in software developed with
predominantly public funds (whether or not the software is copyrighted) and to take only "government
purpose rights" when funding is predominantly but not exclusively private (see pp. 38-39).
2.2.2.7 Surrender the potential unlimited rights claim to software documentation that might
be In a manual or that might be construed as Instrt -1onal material for Installation, operation,
maintenance or training purposes.
Under the SDRC, the DoD acquires unlimited rights in manuals or instructional materials prepared or
required to be delivered under a government contract for Installation, operation, maintenance or train-

73

%ing purposes, even though such manuals may have been developed at private expense and are not
in the public domain.
Although privately developed other-than-commercial-software may receive restricted rights treatment,
manuals or instructional materials for such software, even though they contain proprietary Information, would seem to be governed by the unlimited rights provision. This creates a significant disincentive to do business with DoD and could lead to firms providing the government with no more than the
barest minimum of documentation needed to meet contract requirements (see pp. 23-24).
2.3.2.8 Examine the need for "unlimited rights" as opposed to "rights fer government
pUrp•oses".
In .Accordance with the regulatory policy that DoD shall acquire only such rights to use, duplicate and
disclose software developed at private expense as are necessary to meet government needs, consideration should be given to restructuring the unlimited rights policy to afford the government unlimited rights only where they are truly needed (see pp. 38-43).

2.3.3. Policy as to Privately Funded Software
2.3.3.1 Add to the minimum restricted rights the government obtains In privately developed
software the right to make a copy for reverse engineering purposes If necessary to make
modifications.

':

The restricted rights provisions of the SDRC seems to limit the government's right to copy software to
archival or back-up purposes. Although the minimum rights do include the right to modify the software, if insufficient documentation has been obtained or it is not possible to have the original contractor modify the software, the government may attempt to reverse engineer it. It is not clear under the
regulations or the copyright law whether the modification right includes the right to make a copy for
reverse engineering purposes. In light of the potential risks, it would be prudent for DoD to dearly
state that it has this right. (p. 55.)
2.3.3.2 Develop a standard policy for acquiring privately developed software for local area
networks.
Since local areas networks which share software are becoming more commonplace within DoD, the
regulations should provide guidance about acquiring software intended for use in such networks. (p.
27-28.)
2.3.3.3 Clearly establish the status of restricted rights software which the government has
modified.
When the government modifies privately developed software in which it has restricted rights, the
effect of that modification appears to vary, depending on whether the software is subject to commercial or other-than-commercial restricted rights. The SDRC provides that as to commercial software,
"unmodified portions shall remain subject to these restrictions." However, modifications to other than
commercial software are governed by another subsection of the clause, which provides that "those

16

74

%portions of the derivative software incorporating restricted rfghts software are subject to tie sanw
restricted rights." This apparentIy inconsistent treatment of modifications to restricted rights software
Is extremely confusing and needs to be clarified. (p.66-5.)
The ambiguity of the DoD regulations about ownership rights and restrictions as to software modifications may mean that ifthe original software Is protected by copyright law, ItIs copyright law that will fill
In the gaps. Since modifications are derivative works, a host of copyright issues could arise which
could substantially inhibit the government's use of the software to its maximum potential. (Chapter 4.)
2.3.3.4 Consider eliminating the two different sets of restricted rights for commercial and
other-than-commercial software developed at private expense.
As noted above, the SDRC provides for two different sets of restricted rights for commercial and
other-than-commercial software. There appears to be no clear rationale for this differential treatment
and for the correstonding differential treatment of documentation. Moreover, neither the regulation
nor policy provision provide any clear guidance as to when a piece of software qualifies for ormmercial or other-than-commercial treatment.
The resulting confusion and amoiguity can be avoided by establishing a "floor of minimum rights
which the government must have and then allowing arrangements between the "floor" of minimum
rights and the "ceiling" of unlimited rights to be negotiated as the government's needs require
(.,ee pp. 26-27).
2.3.3.5 If DoD chooses to retain the distinction between commercial and other-thancommercial software, eliminate the potential unlimited rights claim In privately developed
other-than-commercial software as to which no separate license agreement has bsen
negotiated.

4'

When other-than-commercial software is being procured, the SDRC stipulates that a separate license
agreement containing the applicable restrictions is to be negotiated and made a part of the government contract, (so long as the government obtains, at a minimum, the four minimum restricted rights
set forth in the clause). When a firm provides privately developed software to DoD but has not
negotiated a separate licensing agreement, an issue arises as to whether the government would get
unlimited rights in the software or only the four minimum restricted rights. The existence of such a
0potential "booby trap" in the regulations could be enough to dissuade the smaller, "high tech" companies from doing business with DoD with the result that the latest innovative software could be unavailable (see pp. 21-23). The SDRC should be revised to make clear that the govemment will have only
the four standard minimum rights in privately developed other-than-commercial software when no
separate licensing agreement is negotiated.
2.3.3.6 Treat privately developed software documentation as subject to the same restrictions
as the machine readable code.
The SDRC treats commercial computer software and its documentation In a manner consistent with
industry practice by providing that both machine readable code and documentation will be governed
by the same set of restricted rights.
17
75

A*A4AI

F

%Of

In contrast, documentation for other-than-commercial software is not aubject to the same set of
restricted rights as the machine readable code but is instead acquired by the government with 1lmited
rights. This gives the government the right to use, disclose and duplicate the documentation throughout the government. Subjecting other than commercial documentation to the broader limited rights
policy not only causes confusion but deters many software firms from selling rights in their most
valuable technology to DoD. (p. 26-27.)
2.3.3.7 Allow contractors to retain the privately developed status for software when only
minor modifications are made to tailor It for government use.
Under the DoD policy, if a company has developed a piece of software wholly at private expense, and
thin under a government procurement contract, makes some minor modilfications to tailor it for intended government use, the company would forfeit restricted rights status for the delivered software if
DoD funds subsidized the modification. This policy deviates from standard commercial practice, and
is viewed by many software firms as inequitable.
Consideration should be given to adopting the proposed FAR's more flexible approach which allows
contractors to retain the privately developed status for their software when only minor modifications
are made for the government (see pp. 25-26).
2.3.3.8 Consideration should be given to restructuring the software procurement process so
as to allow the government the flexibility to take less than the current minimum restricted
rights In software and less than limited rights In documentation In certain situstions.
In some situations iftmay be in the government's best interests to have the flexibility to acquire fewer
rights in privately developed software than the current SDRC permits in exchange for certain concessions from the contractor. -This built-in flexibility could allow the DoD to satisfy a more pressing need
such as:
a) the need to get a warranty on the software which may not be possible unless the government
agrees te permit the developer to perform all the maintenance work (Chapter 11);
b) the need to create an escrow arrangement to obtain access to privately developed source code
that the software firm would otherwise not provide at reasonable cost to the government (see pp.
52-53); and
c) the need to get acce ss to software tools and/or CAD/CAM programs (see pp..50-51, Chapter 10).
2.3.3.9 Rename the pioposed "license rights" provision of the proposed SDRC, If a "fixed
expiration" option Is to e preserved.
The "license rights" co.-cept as originally conceived by the OSD Study Group was to enable the
government to require its contractors to license competitors to use their proprietary data in competitive re-procurement (or maintenance) situations. However, the "license rights" option proposed by
the DoD FAR Supplement appears to focus on obtaining expirations for restrictive legends. "License
rights' is a misnomer for this set of rights, particularly in view of the fact that the SBIR provisions

18

76

%refleci a very different "license rights" policy. Give the new policy a better name, perhaps "fixed
expiration rights," so that people won't get confused. Ixis questionable whether this new option will
be acceptable to industry which can always elect limited or restricted rights protection for its valuable
technologies (see pp. 32-35).

3. Conclusion
it is important to observe that the problems which DoD is experiencing with its software acquisition
policy are not unique to the government. The problems are being experienced industry-wide, and are
due in large part to the unique nature of software and to the lag between the ability to conceptualize
software as a product and the dbvelopment of the end product. The DoD, as the major single
consumer of software, is in a unique and enviable po•;tion to address the difficulties being encoun•ared within the software industry, and to place itself on the leading edge of tie efhort to bring acquisition and licensing practices in line with the technical and economic realities of software development.
By taking this leadership role, the DoD could do much to help maintain tho U.S. lead in software
technology in the world.

References
[11

Par la Samuelson.
Coj ,ments on the ProposedDefense and FederalAcqv isition Regulations.
Technical Repori SEI-86-TM-2, Softw3re Engineering Institute, Pittsburgh, PA 15213, 1986.

[2]

Pamela Samuielson.
Toward a Re,orin of the Defense DepartmentSoftware Acquisition Policy.
Technical Report CMU/SEI-86-TR-1, Software Engineering Institute, Pitts.burgh, PA 15213,

" .1986.

19
77

%Appendix A7 - A Proposal for an Ada Software Module Market
An Ada Software Module Market enterprise could perhaps operate and become viable
on the following basis:
1. The ASOMM would be established for the purpose of supporting Ada through disseminating Ada modules. It would also accept Ada support tools written in other
languages.
2. The ASOMM would set standards for module catalog description that specify precisely
the portability properties (e.g., "compiles and operates in Unix 4.2 bsd environments,
including these which we have tested: SUN, VAX11"). It might also set standards
for other description attributes - accuracy, speed, function. It would set standards
for the form of source code and documentation and of test cases and test drivers.
3. The owners of modules would set the prices for their ,iferings, but ASOMM would
have a uniform set of terms and conditions, so that prospective users would have
minimum paperwork.
4. ASOMM would handle all marketing, distribution, and licensing, charging a substantial (but perhaps sliding) commission on revenues.
5. ASOMM would not itself undertake module development, enhancement, documentation, repair, support, validation, certification, or warranty. It would be a mail-order
software dealer.
6. All modules would include full copyrighted source, except perhaps for security submodules, which might be object-only.
7. The standard terms and conditions would include at least four kinds of licenses, at
different prices:
* One copy, limited-period trial use, price refundable except for rental fee.
9 Per machine/copy, unlimited use
• Site linceses, at least for personal computers and workstations
* Per incorporation as a component in a larger product, with free sub-licensing
rights. This would enable an incorporator to do a one-time transaction and not
undertake a long-run paperwork burden.
8. The standard terms and conditions would include different levels of support for
licensed users (but not sub-licensees):
* Fully supported by the owner, perhaps for an annual fee.
9 Supported by the owner on a fixed-fee-per-fix basis.
* Support negotiable with the owner.
* Unsupported. Caveat emptor.
9. ASOMM would maintain lists of licensed users for recall and update purposes.
10. For a modest surcharge, a user could get, along with the module, a list of the other
licensed user, willing to be listed.

78

'*

MWl*

%
\end{document}
